---
title: "trillium_phytoplasma"
author: "Julia Boyle"
date: "2023-05-27"
output: html_document
---
# Field data
```{r}
library(tidyverse)
library(cowplot)
library(patchwork)

trillium<-read.csv("Trillium_Field_Data_Frequency.csv")#just 2023
trilliumY<-read.csv("Trillium_Field_Data - Multiyear.csv")#includes 2018 and 2023
trilliumY
trilliumY$Year<-as.factor(trilliumY$Year)
trilliumY
trillium_long<-trilliumY %>% 
  pivot_longer(cols=8:13,
  names_to = c("Status", "Hemiptera_presence"),
  names_sep = "_",
  values_to = "count"
)

trillium_long<-trillium_long %>% mutate(Infection= ifelse(Status=="Infec", "1", "0"))
str(trillium_long)
trillium_long$Infection<-as.integer(trillium_long$Infection)
trillium_long$Year<-as.factor(trillium_long$Year)
trillium_long_adults<-trillium_long %>% 
  filter(Status!="Young")
trillium_long
```

```{r}
hist(log(trilliumY$Freq_infection_of_adults))
hist(trilliumY$Freq_hemiptera_total)
hist(trilliumY$Total_trilliums)
summary(lm(log(Freq_infection_of_adults+1)~Freq_hemiptera_adults*Total_adult_trilliums, data=trillium))#only gets at 2023 data, sig hemiptera and interaction
plot(lm(log(Freq_infection_of_adults+1)~Freq_hemiptera_adults*Total_adult_trilliums, data=trilliumY))

summary(lm(log(Freq_infection_of_adults+1)~Total_adult_trilliums+Year, data=trilliumY)) #uses both years of data but can't include hemiptera, sig density and year
```

```{r}
trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Forest_side))+geom_point()+geom_smooth(method='gam', se=F)+facet_wrap(~Year)

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Forest_side))+geom_point()+facet_wrap(~Year)+theme_cowplot()

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method='gam', se=F)+facet_wrap(~Year)+theme_cowplot()+xlab("Density of flowering trillium in 1m^2")+ylab("Frequency of infected adults")
trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Year))+geom_point()+geom_smooth(method='gam', se=F)


trilliumY %>% filter(Year=='2023') %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults))+geom_point()+xlab("Density of adults in 1m^2")+ylab("Frequency of infected adults")+theme_cowplot()

trilliumY %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm")
```


```{r}
hist(trillium$Freq_hemiptera_adults)
hist(trillium$Freq_infection_of_adult)
trillium %>%
  ggplot(aes(x=Freq_hemiptera_adults, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm")+xlab("Frequency of hemiptera on adults")+ylab("Frequency of infected adults")+theme_cowplot()+ylim(0,0.23)
```
```{r}
trillium %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_hemiptera_adults))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Density of flowering trilliums")+ ylab("Frequency of hemiptera on flowering trilliums")
trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_hemiptera_total))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Density of all trilliums no./m^2")+ ylab("Frequency of hemiptera on all trilliums")
#in very dense patches, hemiptera cannot colonize all the plants, meaning many will remain aphid-free and disease free
#in low density patches, almost all plants may get hemiptera, and so the likelihood of seeing infection is heightened
#predator satiation? 
#additionally, there is evidence sparse plant spacing can increase aphid population density (Chung et al. 1980) https://krex.k-state.edu/bitstream/handle/2097/35613/rec123-chung-1980.pdf?sequence=1&isAllowed=y
#survivorship bias? fewer surviving 
#defensive plant compounds released with herb, more effective at high density
```

```{r}
trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm")
trillium %>% 
  ggplot(aes(x=Freq_infection_of_adults, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Infection frequency of adults")+ylab("Young:Reproductive ratio")
```


# iNaturalist data
```{r}
library(maps)
library(mapdata)
library(ggmap)
library(RColorBrewer)
#library(maptools)
#library(rnaturalearth)
#library(rnaturalearthdata)
#library(rnaturalearthhires)
#library(raster)
inatinfec<-read.csv("iNat_infectrilliums_2023_07_5.csv")
inat<-read.csv("iNat_alltrilliums_2023_07_05.csv")
head(inat)

#time series
inat20s<-inat %>% filter(observed_on>2020-1-1, observed_on<2030-1-1)
inat10s<-inat %>% filter(observed_on>2010-1-1, observed_on<2020-1-1)
inat00s<-inat %>% filter(observed_on>2000-1-1, observed_on<2010-1-1)
inat90s<-inat %>% filter(observed_on>1990-1-1, observed_on<2000-1-1)
inat80sandbelow<-inat %>% filter(observed_on<1990-1-1)

#infected time series
inat20s_infec<-inat20s %>% filter( field.phytoplasma.infection.=="Yes")
inat10s_infec<-inat10s %>% filter( field.phytoplasma.infection.=="Yes")
inat00s_infec<-inat00s %>% filter( field.phytoplasma.infection.=="Yes")
inat90s_infec<-inat90s %>% filter( field.phytoplasma.infection.=="Yes")
inat80sandbelow_infec<-inat80sandbelow %>% filter( field.phytoplasma.infection.=="Yes")
```

But maybe also try having size of bubbles indicating number of infected https://stackoverflow.com/questions/60021868/producing-heat-map-over-geo-locations-in-r 

TO DO:
-Turn lat long into equal area projection (eg. 1km by 1km squares), so we have gridded cells. Look at proportion infected to total observations in each cell.
-Look at year by year (or binned time periods) proportion infected individuals 
-Add user ID as random effect or limit number of observations per user? Or see what the max number of infected observations there are by user ID to see if anyone was specifically seeking them out or not, useful to report
-occupancy model for space: different number of samples per grid, accounting for them



```{r}
#http://mbontrager.org/blog/2016/08/01/Maps-in-R
infectedpoints <- filter(inat, field.phytoplasma.infection.=="Yes")
normalpoints <- filter(inat, field.phytoplasma.infection.=="")
infcolour <- brewer.pal(6, "Paired")[5]
normcolour <- brewer.pal(6, "Paired")[3]

usa <- map_data("usa")
canada <- map_data("worldHires","Canada")
NAmap <- ggplot() + geom_polygon(data = usa, 
                                 aes(x=long, y = lat, group = group), 
                                 fill = "white", 
                                 color="black") +
    geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
                 fill = "white", color="black") +
    coord_fixed(xlim = c(-95, -70),  ylim = c(33, 50), ratio = 1.2)
NAmap

range<-NAmap+ geom_point(data=inat, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 1967-2023")
infections<-NAmap+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 1989-2023")
range+infections
```
Time series
```{r}
#2020 and above
NAmap+ geom_point(data=inat20s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 2020-2023")
NAmap+ geom_point(data=inat20s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 2020-2023")

#2010-2020
NAmap+ geom_point(data=inat10s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 2010-2020")
NAmap+ geom_point(data=inat10s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 2010-2020")

#2000-2010
NAmap+ geom_point(data=inat00s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 2000-2010")
NAmap+ geom_point(data=inat00s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 2000-2010")

#1990-2000
NAmap+ geom_point(data=inat90s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 1990-2000")
NAmap+ geom_point(data=inat90s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 1990-2000")

#1980s and below
NAmap+ geom_point(data=inat80sandbelow, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 1980s and below")
NAmap+ geom_point(data=inat80sandbelow_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 1980s and below")
```


```{r}
#https://axelhodler.medium.com/creating-a-heat-map-from-coordinates-using-r-780db4901075
map_bounds <- c(left = -90, bottom = 30, right = -70, top = 50)
coords.map <- get_stamenmap(map_bounds, zoom = 7, maptype = "toner-lite")
coords.map <- ggmap(coords.map, extent="device", legend="none")

coords.map.infecscale <- coords.map + stat_density2d(data=inatinfec,  aes(x=longitude, y=latitude, fill=..level.., alpha=..level..), geom="polygon")+scale_fill_gradientn(colours=rev(brewer.pal(7, "Spectral")))+ theme_bw()+xlab("Longitude")+ylab("Latitude")

coords.map+ geom_point(data=inatinfec,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")

coords.map.infecscale+ geom_point(data=inatinfec,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)
coords.map.infecscale

coords.map.scale <- coords.map + stat_density2d(data=inat,  aes(x=longitude, y=latitude, fill=..level.., alpha=..level..), geom="polygon")+scale_fill_gradientn(colours=rev(brewer.pal(7, "Spectral")))+ theme_bw()+xlab("Longitude")+ylab("Latitude")

coords.map+ geom_point(data=inat,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")

coords.map.scale+ geom_point(data=inat,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)
coords.map.scale
```


