---
title: "trillium_phytoplasma"
author: "Julia Boyle"
date: "2023-05-27"
output: html_document
---
# Field data

## Frequency of infection and hemiptera
look at prevalence vs presence more? 

```{r}
library(tidyverse)
library(cowplot)
library(patchwork)
library(car)
library(GLMMadaptive)
library(pscl)

trillium<-read.csv("Trillium_Field_Data_Frequency.csv")#just 2023
trillium<-trillium %>% mutate(Total_hemiptera_trillium=Young_hemiptera+Health_hemiptera+Infec_hemiptera)
trilliumY<-read.csv("Trillium_Field_Data - Multiyear.csv")#includes 2018 and 2023
trilliumY
trilliumY$Year<-as.factor(trilliumY$Year)
trilliumY
trillium_long<-trilliumY %>% 
  pivot_longer(cols=8:13,
  names_to = c("Status", "Hemiptera_presence"),
  names_sep = "_",
  values_to = "count"
)

trillium_long<-trillium_long %>% mutate(Infection= ifelse(Status=="Infec", "1", "0"))
str(trillium_long)
trillium_long$Infection<-as.integer(trillium_long$Infection)
trillium_long$Year<-as.factor(trillium_long$Year)
trillium_long_adults<-trillium_long %>% 
  filter(Status!="Young")
trillium_long
```

```{r}
hist(log(trilliumY$Freq_infection_of_adults+1))
hist(trilliumY$Freq_infection_of_adults)
hist(trilliumY$Freq_hemiptera_total)
hist(trilliumY$Total_trilliums)
```
Zero-inflated poisson models
```{r}
#2018 and 2023
##with random effects
head(trilliumY)
str(trilliumY)
MM4<-mixed_model(Total_infec_trilliums~Total_adult_trilliums+Year, random = ~ 1 | Population, data = trilliumY, family = zi.poisson(), zi_fixed = ~ Total_adult_trilliums+Year, zi_random=~1 |Population,control=list(iter_EM = 0))
summary(MM4) #loglik -254.7688. count model: density p=0.069. zero model:	 higher density increases likelihood of finding 0s p=0.01, higher year decreases chance of finding 0s=0.0388
##without random effects
summary(zeroinfl(Total_infec_trilliums~Total_adult_trilliums+Year, data=trilliumY)) #loglik -272.1, basically same results as above

#2023 only
##with random effects
head(trillium)
MM2<-mixed_model(Total_infec_trilliums~Freq_hemiptera_adults+Total_adult_trilliums, random = ~ 1 | Population, data = trillium, family = zi.poisson(), zi_fixed = ~ Freq_hemiptera_adults+Total_adult_trilliums, zi_random=~1 |Population,control=list(iter_EM = 0))
summary(MM2) #loglik -83.28, freq hemip 0.057 in count model 

##without random effects
summary(zeroinfl(Total_infec_trilliums~Freq_hemiptera_adults+Total_adult_trilliums, data=trillium)) #log lik -82.6, sig more likely to get zero infection in high adult density , marginal hemip in count model

#correlations
cor(trillium$Total_trilliums, trillium$Freq_hemiptera_total) #-0.475946
cor(trillium$Freq_infection_of_adults, trillium$Freq_hemiptera_adults) #0.3341377
cor(trillium$Freq_infection_of_adults, trillium$Young_to_repro_ratio) #0.1034085
```

```{r}
trillium %>% ggplot(aes(x=Total_trilliums, y=Total_infec_trilliums, color=Freq_hemiptera_adults))+geom_point()+theme_cowplot()+labs(x="Trillium (no.)", y="Symptomatic trillium (no.)")
trillium %>% ggplot(aes(x=Total_adult_trilliums, y=Total_infec_trilliums, color=Freq_hemiptera_adults))+geom_point()+theme_cowplot()+labs(x="Flowering trillium (no.)", y="Symptomatic trillium (no.)")+scale_color_continuous("Freq. of hemiptera on flowering trillium",low = "blue", high = "hotpink")
```
```{r}
trilliumY %>% 
  ggplot(aes(x=Population, y=Freq_infection_of_adults, colour=Population))+geom_boxplot()+facet_wrap(~Forest_side)

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Forest_side))+geom_point()+geom_smooth(method='gam', se=F)+facet_wrap(~Year)

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Forest_side))+geom_point()+facet_wrap(~Year)+theme_cowplot()

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method='gam', se=F)+facet_wrap(~Year)+theme_cowplot()+xlab("Density of flowering trillium in 1m^2")+ylab("Frequency of infected adults")
trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Year))+geom_point()+geom_smooth(method='gam', se=F)


trilliumY %>% filter(Year=='2023') %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults))+geom_point()+xlab("Density of adults in 1m^2")+ylab("Frequency of infected adults")+theme_cowplot()

trilliumY %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm")
```

```{r}
hist(trillium$Freq_hemiptera_adults)
hist(trillium$Freq_infection_of_adult)
trillium %>%
  ggplot(aes(x=Freq_hemiptera_adults, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm")+xlab("Frequency of hemiptera on adults")+ylab("Frequency of infected adults")+theme_cowplot()+ylim(0,0.23)
```
```{r}
trillium %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_hemiptera_adults))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Density of flowering trilliums")+ ylab("Frequency of hemiptera on flowering trilliums")
trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_hemiptera_total))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Density of all trilliums no./m^2")+ ylab("Frequency of hemiptera on all trilliums")
#in very dense patches, hemiptera cannot colonize all the plants, meaning many will remain aphid-free and disease free
#in low density patches, almost all plants may get hemiptera, and so the likelihood of seeing infection is heightened
#predator satiation? 
#additionally, there is evidence sparse plant spacing can increase aphid population density (Chung et al. 1980) https://krex.k-state.edu/bitstream/handle/2097/35613/rec123-chung-1980.pdf?sequence=1&isAllowed=y
#survivorship bias? fewer surviving 
#defensive plant compounds released with herb, more effective at high density
```

```{r}
trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm")
trillium %>% 
  ggplot(aes(x=Freq_infection_of_adults, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Infection frequency of adults")+ylab("Young:Reproductive ratio")
```
Figure
```{r}
fig2A<-trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method='gam', se=F, color="yellowgreen")+facet_wrap(~Year)+theme_cowplot()+xlab("Density of flowering trillium in 1m^2")+ylab("Frequency of infected adults")
fig2B<-trillium %>%
  ggplot(aes(x=Freq_hemiptera_adults, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm", color="yellowgreen")+xlab("Frequency of hemiptera on flowering trillium")+ylab("Frequency of infected flowering trillium")+theme_cowplot()+ylim(0,0.23)
fig2C<-trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_hemiptera_total))+geom_point()+geom_smooth(method="lm", color="yellowgreen")+theme_cowplot()+xlab("Density of all trilliums no./m^2")+ ylab("Frequency of hemiptera on all trilliums")
fig2D<-trillium %>% 
  ggplot(aes(x=Freq_infection_of_adults, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm", color="yellowgreen")+theme_cowplot()+xlab("Infection frequency of flowering trillium")+ylab("Non-flowering:Flowering ratio")

plot_grid(fig2A,fig2B,fig2C, fig2D, labels=c("A", "B", "C", "D")) #save 10x12
plot_grid(fig2A,fig2B,fig2C, labels=c("A", "B", "C"), ncol=1)
fig2Alabelled<-fig2A+plot_annotation(tag_levels = list(c("A")))
fig2BC<-fig2B/fig2C+plot_annotation(tag_levels = list(c("B", "C")))+ plot_layout(ncol=1)
plot_grid(fig2Alabelled,fig2BC, ncol=1)
fig2Alabelled+fig2BC
```


## Trait data
```{r}
traits<-read.csv("Trillium_Field_Data - Leaf and Height Phenotypes.csv")
str(traits)
traits<-traits %>% mutate(Infection_chr=ifelse(Infection=="0", "Asymptomatic", "Symptomatic"))
```
```{r}
#Height
figheight<-traits %>% ggplot(aes(x=Infection_chr, y=Height_cm))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Height (cm)")
traits %>% ggplot(aes(x=Infection, y=Height_cm))+geom_point()+geom_smooth()+theme_cowplot()
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Height_cm))
1-(23.34688/27.61000	)
27.61-23.34688

#leaf length
traits %>% ggplot(aes(x=Infection_chr, y=Leaf_length_cm))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Leaf length (cm)")

#sepal length
figsepal<-traits %>% ggplot(aes(x=Infection_chr, y=Sepal_length_cm))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Sepal length (cm)")
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Sepal_length_cm))
1-(3.460667/4.820938)

#leaf and sepal ratio
figratio<-traits %>% ggplot(aes(x=Infection_chr, y=Sepal_to_leaf_ratio))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Sepal:Leaf length ratio")
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Sepal_to_leaf_ratio))

#figure
plot_grid(figheight,figsepal, figratio, labels=c("A", "B", "C"))
```
```{r}
hist(traits$Height_cm)
summary(lmer(Height_cm~Infection_chr+(1|Plot/Population), data=traits))
Anova(lmer(Height_cm~Infection_chr+(1|Plot/Population), data=traits), type=3) #p=0.003096

hist(traits$Leaf_length_cm)
Anova(lmer(Leaf_length_cm~Infection_chr+(1|Plot/Population), data=traits), type=3) #not sig

hist(log(traits$Sepal_length_cm))
shapiro.test(log(traits$Sepal_length_cm))
summary(lmer(log(Sepal_length_cm)~Infection_chr+(1|Plot/Population), data=traits))
Anova(lmer(log(Sepal_length_cm)~Infection_chr+(1|Plot/Population), data=traits), type=3) #p=2.342e-07
```


## Sticky trap data
```{r}
sticky<-read.csv("Trillium_Field_Data - Sticky traps.csv")
head(sticky)
str(sticky)
sticky$Adult_Trillium_density<-factor(sticky$Adult_Trillium_density, levels=c("Low", "Mid", "High"))
sticky_noout<-sticky %>% filter(Outlier=="No")
```

```{r}
sticky_noout %>% ggplot(aes(x=Adult_Trillium_density, y=Num_sapsuckers_on_trap))+geom_boxplot()
sticky_noout %>% ggplot(aes(x=Adult_Trillium_density, y=Total_insects_on_trap))+geom_boxplot()
sticky_noout %>% ggplot(aes(x=Forest, y=Num_sapsuckers_on_trap, fill=Adult_Trillium_density))+geom_boxplot()+theme_cowplot()+scale_fill_grey(name="Density of flowering trillium", start=0.5, end=1)+labs(x = "Forest population",y = "Hemiptera caught (no.)")+scale_x_discrete(labels=c("Old growth", "Secondary"))
sticky_noout %>% ggplot(aes(x=Forest, y=Total_insects_on_trap, color=Adult_Trillium_density))+geom_boxplot()
#this mainly shows that old forest had more bugs than secondary forest, and that intermediate trillium density plots tended to have the most insects

sticky_noout %>% ggplot(aes(x=Num_sapsuckers_on_trap, y=Freq_infection_of_adults, color=Adult_Trillium_density))+geom_point()+theme_cowplot()
sticky_noout %>% ggplot(aes(x=Num_sapsuckers_on_trap, y=Freq_hemiptera_total, color=Adult_Trillium_density))+geom_point()+theme_cowplot()
```

# iNaturalist data
```{r}
library(maps)
library(mapdata)
library(ggmap)
library(RColorBrewer)
library(raster)
library(scales)
library(CoordinateCleaner)
#library(maptools)
library(rnaturalearth)
library(rnaturalearthdata)
#library(geodata)
library(sf)
library(plyr)
library(data.table)
#library(tidyterra)
#library(rnaturalearthhires)
#library(raster)
inat<-read.csv("iNat_alltrilliums_2023_07_05.csv")
head(inat)
inat<-inat %>% mutate(country_code=ifelse(grepl("Canada", place_country_name), "CAN", "USA"))
inat <- inat %>%
  filter(!is.na(longitude)) %>%
  filter(!is.na(latitude))

inat_clean <- clean_coordinates(x = inat, 
                           lon = "longitude", 
                           lat = "latitude",
                           countries = "country_code",
                           species = "scientific_name",
                           tests = c("centroids",
                                    "equal", "zeros", "institutions", "duplicates")) 
inat_clean=inat_clean %>% filter(.summary==TRUE) #only include observations that have no suspect issues

set.seed(1)
overrepresented_users<-inat_clean %>% group_by(user_login) %>% tally() %>% filter(n>15)
overrepresented_samples<-subset(inat_clean, user_login %in% overrepresented_users$user_login)
overrepresented_subset<-plyr::ddply(overrepresented_samples,.(user_login),function(x) x[sample(nrow(x),15),])#make the max number of observations per user 15 
normalrepresented_users<-inat_clean %>% group_by(user_login) %>% tally() %>% filter(n<=15)
normalrepresented_samples<-subset(inat_clean, user_login %in% normalrepresented_users$user_login)
inat_clean_corrected<-rbind(overrepresented_subset,normalrepresented_samples)
inat_clean_corrected
inat_clean_corrected_tally<-inat_clean_corrected %>% group_by(user_login) %>% tally()
mean(inat_clean_corrected_tally$n)
median(inat_clean_corrected_tally$n)

inatinfec<-inat_clean_corrected %>% filter( field.phytoplasma.infection.=="Yes") #make another dataset of infected points
inatinfec %>% group_by(user_login) %>% tally() #no user contributed more than 6 infected observations

#time series
inat20s<-inat_clean_corrected %>% filter(observed_on>2020-1-1, observed_on<2030-1-1)
inat10s<-inat_clean_corrected %>% filter(observed_on>2010-1-1, observed_on<2020-1-1)
inat00s<-inat_clean_corrected %>% filter(observed_on>2000-1-1, observed_on<2010-1-1)
inat90s<-inat_clean_corrected %>% filter(observed_on>1990-1-1, observed_on<2000-1-1)
inat80sandbelow<-inat_clean_corrected %>% filter(observed_on<1990-1-1)

#infected time series
inat20s_infec<-inat20s %>% filter( field.phytoplasma.infection.=="Yes")
inat10s_infec<-inat10s %>% filter( field.phytoplasma.infection.=="Yes")
inat00s_infec<-inat00s %>% filter( field.phytoplasma.infection.=="Yes")
inat90s_infec<-inat90s %>% filter( field.phytoplasma.infection.=="Yes")
inat80sandbelow_infec<-inat80sandbelow %>% filter( field.phytoplasma.infection.=="Yes")
```

But maybe also try having size of bubbles indicating number of infected https://stackoverflow.com/questions/60021868/producing-heat-map-over-geo-locations-in-r 

TO DO:
-Turn lat long into equal area projection (eg. 1km by 1km squares), so we have gridded cells. Look at proportion infected to total observations in each cell.
-occupancy model for space: different number of samples per grid, accounting for them
-see what predicts infection?? worldclim data or crop cover? human population density? trillium density? Vector density? ****

Main maps
```{r}
#http://mbontrager.org/blog/2016/08/01/Maps-in-R
infectedpoints <- filter(inat_clean_corrected, field.phytoplasma.infection.=="Yes")
normalpoints <- filter(inat_clean_corrected, field.phytoplasma.infection.=="")
infcolour <- brewer.pal(6, "Paired")[5]
normcolour <- brewer.pal(6, "Paired")[3]

usa <- map_data("state")
canada <- map_data("worldHires","Canada")

NAmap <- ggplot() + geom_polygon(data = usa, 
                                 aes(x=long, y = lat, group = group), 
                                 fill = "white", 
                                 color="black") +
    geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
                 fill = "white", color="black") +
    coord_fixed(xlim = c(-95, -70),  ylim = c(33, 50), ratio = 1.2)
NAmap

range<-NAmap+ geom_point(data=inat_clean_corrected, aes(x=longitude, y=latitude), fill="hotpink",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "*Trillium grandiflorum* observations") +theme(title = ggtext::element_markdown())
infections<-NAmap+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), fill="yellowgreen",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "Infected *T. grandiflorum* observations") +theme(title = ggtext::element_markdown())
rangemap<-plot_grid(range,infections, labels=c("A", "B"))
ggsave2("phytoplasma_rangemap.png", rangemap, dpi=600, width=10, height=8)
```

Time series
```{r}
#2020 and above
NAmap+ geom_point(data=inat20s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 2020-2023")
NAmap+ geom_point(data=inat20s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 2020-2023")

#2010-2020
NAmap+ geom_point(data=inat10s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 2010-2020")
NAmap+ geom_point(data=inat10s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 2010-2020")

#2000-2010
NAmap+ geom_point(data=inat00s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 2000-2010")
NAmap+ geom_point(data=inat00s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 2000-2010")

#1990-2000
NAmap+ geom_point(data=inat90s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 1990-2000")
NAmap+ geom_point(data=inat90s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 1990-2000")

#1980s and below
NAmap+ geom_point(data=inat80sandbelow, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 1980s and below")
NAmap+ geom_point(data=inat80sandbelow_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 1980s and below")
```


```{r}
#https://axelhodler.medium.com/creating-a-heat-map-from-coordinates-using-r-780db4901075
map_bounds <- c(left = -90, bottom = 30, right = -70, top = 50)
coords.map <- get_stamenmap(map_bounds, zoom = 7, maptype = "toner-lite")
coords.map <- ggmap(coords.map, extent="device", legend="none")

coords.map.infecscale <- coords.map + stat_density2d(data=inatinfec,  aes(x=longitude, y=latitude, fill=..level.., alpha=..level..), geom="polygon")+scale_fill_gradientn(colours=rev(brewer.pal(7, "Spectral")))+ theme_bw()+xlab("Longitude")+ylab("Latitude")

coords.map+ geom_point(data=inatinfec,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")

coords.map.infecscale+ geom_point(data=inatinfec,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)
coords.map.infecscale

coords.map.scale <- coords.map + stat_density2d(data=inat,  aes(x=longitude, y=latitude, fill=..level.., alpha=..level..), geom="polygon")+scale_fill_gradientn(colours=rev(brewer.pal(7, "Spectral")))+ theme_bw()+xlab("Longitude")+ylab("Latitude")

coords.map+ geom_point(data=inat,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")

coords.map.scale+ geom_point(data=inat,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)
coords.map.scale
```


