---
title: "trillium_phytoplasma"
author: "Julia Boyle"
date: "2023-05-27"
output: html_document
---
# Field data

## Frequency of infection and hemiptera
look at prevalence vs presence more? 

```{r}
library(tidyverse)
library(cowplot)
library(patchwork)
library(car)
library(GLMMadaptive)
library(pscl)

trillium<-read.csv("Trillium_Field_Data_Frequency.csv")#just 2023
trillium<-trillium %>% mutate(Total_hemiptera_trillium=Young_hemiptera+Health_hemiptera+Infec_hemiptera)
trilliumY<-read.csv("Trillium_Field_Data - Multiyear.csv")#includes 2018 and 2023
trilliumY
trilliumY$Year<-as.factor(trilliumY$Year)
trilliumY
trillium_long<-trilliumY %>% 
  pivot_longer(cols=8:13,
  names_to = c("Status", "Hemiptera_presence"),
  names_sep = "_",
  values_to = "count"
)
trilliumY<-trilliumY %>% mutate(Infection_presence=ifelse(Freq_infection_of_adults>0,"Yes","No"))

trillium_long<-trillium_long %>% mutate(Infection= ifelse(Status=="Infec", "1", "0"))
str(trillium_long)
trillium_long$Infection<-as.integer(trillium_long$Infection)
trillium_long$Year<-as.factor(trillium_long$Year)
trillium_long_adults<-trillium_long %>% 
  filter(Status!="Young")
trillium_long
```

```{r}
hist(log(trilliumY$Freq_infection_of_adults+1))
hist(trilliumY$Freq_infection_of_adults)
hist(trilliumY$Freq_hemiptera_total)
hist(trilliumY$Total_trilliums)
```
Zero-inflated poisson models
```{r}
#2018 and 2023
##with random effects
head(trilliumY)
str(trilliumY)
MM4<-mixed_model(Total_infec_trilliums~Total_adult_trilliums+Year, random = ~ 1 | Population, data = trilliumY, family = zi.poisson(), zi_fixed = ~ Total_adult_trilliums+Year, zi_random=~1 |Population,control=list(iter_EM = 0))
summary(MM4) #loglik -254.7688. count model: density p=0.069. zero model:	 higher density increases likelihood of finding 0s p=0.01, higher year decreases chance of finding 0s=0.0388
##without random effects
summary(zeroinfl(Total_infec_trilliums~Total_adult_trilliums+Year, data=trilliumY)) #loglik -272.1, basically same results as above

#2023 only
##with random effects
head(trillium)
MM2<-mixed_model(Total_infec_trilliums~Freq_hemiptera_adults+Total_adult_trilliums, random = ~ 1 | Population, data = trillium, family = zi.poisson(), zi_fixed = ~ Freq_hemiptera_adults+Total_adult_trilliums, zi_random=~1 |Population,control=list(iter_EM = 0))
summary(MM2) #loglik -83.28, freq hemip 0.057 in count model 

##without random effects
summary(zeroinfl(Total_infec_trilliums~Freq_hemiptera_adults+Total_adult_trilliums, data=trillium)) #log lik -82.6, sig more likely to get zero infection in high adult density , marginal hemip in count model

#correlations
cor(trillium$Total_trilliums, trillium$Freq_hemiptera_total) #-0.475946
cor(trillium$Freq_infection_of_adults, trillium$Freq_hemiptera_adults) #0.3341377
cor(trillium$Freq_infection_of_adults, trillium$Young_to_repro_ratio) #0.1034085
```

```{r}
hist(trillium$Freq_hemiptera_adults)
hist(log(trillium$Freq_infection_of_adults))
hist(trillium$Freq_hemiptera_total)
hist(log(trillium$Total_trilliums))
hist(log(trillium$Young_to_repro_ratio))

summary(lm(Freq_infection_of_adults~Freq_hemiptera_adults, data=trillium))

summary(lm(Freq_hemiptera_total~Total_trilliums, data=trillium))

summary(lm(log(Young_to_repro_ratio)~Freq_infection_of_adults, data=trillium))
```


```{r}
trillium %>% ggplot(aes(x=Total_trilliums, y=Total_infec_trilliums, color=Freq_hemiptera_adults))+geom_point()+theme_cowplot()+labs(x="Trillium (no.)", y="Symptomatic trillium (no.)")
trillium %>% ggplot(aes(x=Total_adult_trilliums, y=Total_infec_trilliums, color=Freq_hemiptera_adults))+geom_point()+theme_cowplot()+labs(x="Flowering trillium (no.)", y="Symptomatic trillium (no.)")+scale_color_continuous("Freq. of hemiptera on flowering trillium",low = "blue", high = "hotpink")
```
```{r}
trilliumY %>% 
  ggplot(aes(x=Population, y=Freq_infection_of_adults, colour=Population))+geom_boxplot()+facet_wrap(~Forest_side)

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Forest_side))+geom_point()+geom_smooth(method='gam', se=F)+facet_wrap(~Year)

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Forest_side))+geom_point()+facet_wrap(~Year)+theme_cowplot()

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method='gam', se=F)+facet_wrap(~Year)+theme_cowplot()+xlab("Density of flowering trillium in 1m^2")+ylab("Frequency of infected adults")
trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Year))+geom_point()+geom_smooth(method='gam', se=F)


trilliumY %>% filter(Year=='2023') %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults))+geom_point()+xlab("Density of adults in 1m^2")+ylab("Frequency of infected adults")+theme_cowplot()

trilliumY %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm")
```

```{r}
hist(trillium$Freq_hemiptera_adults)
hist(trillium$Freq_infection_of_adult)
trillium %>%
  ggplot(aes(x=Freq_hemiptera_adults, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm", formula=y~x+0)+xlab("Frequency of hemiptera on adults")+ylab("Frequency of infected adults")+theme_cowplot()#+ylim(0,0.23) #forcing it to pass through 0??
```
```{r}
trillium %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_hemiptera_adults))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Density of flowering trilliums")+ ylab("Frequency of hemiptera on flowering trilliums")
trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_hemiptera_total))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Density of all trilliums no./m^2")+ ylab("Frequency of hemiptera on all trilliums")
#in very dense patches, hemiptera cannot colonize all the plants, meaning many will remain aphid-free and disease free
#in low density patches, almost all plants may get hemiptera, and so the likelihood of seeing infection is heightened
#predator satiation? 
#additionally, there is evidence sparse plant spacing can increase aphid population density (Chung et al. 1980) https://krex.k-state.edu/bitstream/handle/2097/35613/rec123-chung-1980.pdf?sequence=1&isAllowed=y
#survivorship bias? fewer surviving 
#defensive plant compounds released with herb, more effective at high density
```

```{r}
trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm")
trillium %>% 
  ggplot(aes(x=Freq_infection_of_adults, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Infection frequency of adults")+ylab("Young:Reproductive ratio")
```
Figures

```{r}
fig2A<-trilliumY %>% group_by(Total_adult_trilliums, Infection_presence, Year) %>% tally() %>% ggplot(aes(Total_adult_trilliums, fill = Infection_presence, colour = Infection_presence)) +
  geom_density(alpha = 0.1)+facet_grid(~Year)+theme_cowplot()+xlab(expression('Density of flowering trillium (no./1 m'^2* ')'))+ylab("Density of sampled plots")+scale_fill_manual(name="Infection presence",values=c("grey", "darkgreen"))+scale_colour_manual(name="Infection presence",values=c("grey", "darkgreen"))+labs(subtitle = "Year p = 0.038, Trillium density p = 0.010")+ theme(legend.position = "inside",legend.position.inside = c(0.25, 0.8))

fig2B<-trilliumY %>% group_by(Total_adult_trilliums, Total_infec_trilliums, Year) %>% tally() %>% ggplot(aes((Total_adult_trilliums), y=Total_infec_trilliums)) +
  geom_point(alpha=0.5)+facet_grid(~Year)+theme_cowplot()+xlab(expression('Density of flowering trillium (no./1 m'^2* ')'))+ylab("Number of infected trillium")+labs(subtitle = "Year p = 0.706, Trillium density p = 0.069")

cor.test(trillium$Total_trilliums, trillium$Freq_hemiptera_total)
fig2C<-trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_hemiptera_total))+geom_point()+geom_smooth(method="lm", color="darkgreen")+theme_cowplot()+xlab(expression('Density of all trillium (no./1 m'^2* ')'))+ ylab("Proportion of all trillium with hemiptera")+labs(subtitle = "r = -0.476, p < 0.001")

cor.test(trillium$Freq_hemiptera_adults, trillium$Freq_infection_of_adults)
fig2D<-trillium %>%
  ggplot(aes(x=Freq_hemiptera_adults, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm", color="darkgreen")+xlab("Proportion of flowering trillium with hemiptera")+ylab("Proportion of flowering trillium infected")+theme_cowplot()+labs(subtitle = "r = 0.334, p = 0.006")

plot_grid(fig2A,fig2B,fig2C, fig2D, labels=c("A", "B", "C", "D"))  #10 by 10
```

```{r}
cor.test(trillium$Freq_infection_of_adults, trillium$Young_to_repro_ratio)
demography2023<-trillium %>% 
  ggplot(aes(x=Freq_infection_of_adults, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm", color="darkgreen")+theme_cowplot()+xlab("Proportion of flowering trillium infected")+ylab("Non-flowering:flowering ratio")+labs(subtitle = "r = 0.103, p = 0.412")
```

## Trait data
```{r}
traits<-read.csv("Trillium_Field_Data - Leaf and Height Phenotypes.csv")
str(traits)
traits<-traits %>% mutate(Infection_chr=ifelse(Infection=="0", "Asymptomatic", "Symptomatic"))
```
```{r}
#Height
figheight<-traits %>% ggplot(aes(x=Infection_chr, y=Height_cm))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Height (cm)")
traits %>% ggplot(aes(x=Infection, y=Height_cm))+geom_point()+geom_smooth()+theme_cowplot()
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Height_cm))
1-(23.34688/27.61000	)
27.61-23.34688

#leaf length
traits %>% ggplot(aes(x=Infection_chr, y=Leaf_length_cm))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Leaf length (cm)")

#sepal length
figsepal<-traits %>% ggplot(aes(x=Infection_chr, y=Sepal_length_cm))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Sepal length (cm)")
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Sepal_length_cm))
1-(3.460667/4.820938)

#leaf and sepal ratio
figratio<-traits %>% ggplot(aes(x=Infection_chr, y=Sepal_to_leaf_ratio))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Sepal:Leaf length ratio")
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Sepal_to_leaf_ratio))

#figure
plot_grid(figheight,figsepal, figratio, labels=c("A", "B", "C"))
```
```{r}
hist(traits$Height_cm)
summary(lmer(Height_cm~Infection_chr+(1|Plot/Population), data=traits))
Anova(lmer(Height_cm~Infection_chr+(1|Plot/Population), data=traits), type=3) #p=0.003096

hist(traits$Leaf_length_cm)
Anova(lmer(Leaf_length_cm~Infection_chr+(1|Plot/Population), data=traits), type=3) #not sig

hist(log(traits$Sepal_length_cm))
shapiro.test(log(traits$Sepal_length_cm))
summary(lmer(log(Sepal_length_cm)~Infection_chr+(1|Plot/Population), data=traits))
Anova(lmer(log(Sepal_length_cm)~Infection_chr+(1|Plot/Population), data=traits), type=3) #p=2.342e-07
```

## Sticky trap data
```{r}
sticky<-read.csv("Trillium_Field_Data - Sticky traps.csv")
head(sticky)
str(sticky)
sticky$Adult_Trillium_density<-factor(sticky$Adult_Trillium_density, levels=c("Low", "Mid", "High"))
sticky_noout<-sticky %>% filter(Outlier=="No")
```

```{r}
sticky_noout %>% ggplot(aes(x=Adult_Trillium_density, y=Num_sapsuckers_on_trap))+geom_boxplot()
sticky_noout %>% ggplot(aes(x=Adult_Trillium_density, y=Total_insects_on_trap))+geom_boxplot()
sticky_noout %>% ggplot(aes(x=Forest, y=Num_sapsuckers_on_trap, fill=Adult_Trillium_density))+geom_boxplot()+theme_cowplot()+scale_fill_grey(name="Density of flowering trillium", start=0.5, end=1)+labs(x = "Forest population",y = "Hemiptera caught (no.)")+scale_x_discrete(labels=c("Old growth", "Secondary"))
sticky_noout %>% ggplot(aes(x=Forest, y=Total_insects_on_trap, color=Adult_Trillium_density))+geom_boxplot()
#this mainly shows that old forest had more bugs than secondary forest, and that intermediate trillium density plots tended to have the most insects

sticky_noout %>% ggplot(aes(x=Num_sapsuckers_on_trap, y=Freq_infection_of_adults, color=Adult_Trillium_density))+geom_point()+theme_cowplot()
sticky_noout %>% ggplot(aes(x=Num_sapsuckers_on_trap, y=Freq_hemiptera_total, color=Adult_Trillium_density))+geom_point()+theme_cowplot()
```

## Seed counts 2024
```{r}
#actual seed counts only
seed<-read.csv("2024 Trillium seed counts and predicted counts.csv")
head(seed)
seed$seed_count_round<-round(seed$seed_count)
seed<-seed %>% mutate(Full_treatment=paste(Infection_status, Seed_count_real_or_predicted, sep = '_'))

seed_tril_plot<-seed %>% drop_na(seed_count) %>%  ggplot(aes(x=Infection_status, y=seed_count, colour=Seed_count_real_or_predicted))+geom_boxplot()+geom_point( position=position_jitterdodge(jitter.width=0.25, dodge.width=0.8), alpha=0.2)+theme_cowplot()+ylab("Seed count")+xlab("Infection status")+scale_x_discrete(labels=c("Asymptomatic", "Symptomatic"))+scale_colour_manual(values=c("goldenrod", "darkgrey"),name="Seed count", labels=c("Predicted from leaf length", "Real"))+theme(legend.position="top")

seed %>%filter(Paired_data_plants=='Yes'& Seed_count_real_or_predicted=='Real') %>%  drop_na(seed_count)%>% group_by(Infection_status) %>% tally() #only 81 plants had real seed count numbers
seed %>%filter(Paired_data_plants=='Yes') %>% group_by(Infection_status) %>% filter(seed_count==0) %>% tally()
51/81
seed_comparison1<-seed %>% filter(Paired_data_plants=='Yes'&Infection_status=="Symptomatic")
seed_comparison2<-seed %>% filter(Full_treatment=="Symptomatic_Real" | Full_treatment== "Nonsymptomatic_Predicted_from_lf_length")
hist((seed_comparison2$seed_count))
summary(glmer.nb(round(seed_count)~Seed_count_real_or_predicted+(1|pair_number), data=seed_comparison1)) #comparing infected observed vs. expected seed number based on size
summary(glmer.nb(round(seed_count)~Infection_status+(1|pair_number), data=seed_comparison2,   control = glmerControl("bobyqa"))) #comparing the focal paired plants, actual infected and predicted seed count of healthy paired plant
```

```{r}
plot_grid(seed_tril_plot, demography2023, labels=c("A", "B"))
```


# iNaturalist data
```{r}
#install.packages('geodata')

library(maps)
library(mapdata)
library(ggmap)
library(RColorBrewer)
library(raster)
library(scales)
library(CoordinateCleaner)
#library(maptools)
library(rnaturalearth)
library(rnaturalearthdata)
library(geodata)
library(sf)
library(plyr)
library(spData)
library(spDataLarge)
library(tidyterra)
library(terra)
#library(rnaturalearthhires)
library(tmap) 
library(predicts)
```

```{r}
#inat<-read.csv("iNat_alltrilliums_2023_07_05.csv") #only up until 2023
inat<-read.csv("iNat_alltrilliums_2025_03_14_539340.csv") #version with 2024 data
head(inat)

inat_cleaned = inat %>% 
  # Adding positional accuracy filter to see if that helps deal with observations in lakes
  filter(!is.na(latitude), !is.na(longitude), !is.na(public_positional_accuracy)) %>% 
  filter(public_positional_accuracy < 10000) %>% 
  dplyr::mutate(infected = `field.phytoplasma.infection.`) %>% 
  mutate(infected = if_else(is.na(infected), "No", infected)) %>% 
  # Drop european and puerto rico observations
  filter(latitude > 20, longitude < -65) %>% 
  # Group by user and subset to 15 observations per user
  group_by(user_login) %>% 
  slice_sample(n = 15, replace = FALSE)

inat_clean <- clean_coordinates(x = inat_cleaned, 
                           lon = "longitude", 
                           lat = "latitude",
                           species = "scientific_name",
                           tests = c("centroids",
                                    "equal", "zeros","institutions", "duplicates")) 
inat_clean=inat_clean %>% filter(.summary==TRUE) #only include observations that have no suspect issues
inat_clean=inat_clean %>% mutate(infected=ifelse(grepl("Yes", infected), "1", "0"))
inat_clean$infected

inat%>% filter( field.phytoplasma.infection.=="Yes") #845
inat_cleaned%>% filter( field.phytoplasma.infection.=="Yes") #550
inat_clean%>% filter( field.phytoplasma.infection.=="Yes") #526 after filtering out stuff with clean_coordinates
```
```{r}
set.seed(1)
overrepresented_users<-inat_clean %>% group_by(user_login) %>% tally() %>% filter(n>15)
overrepresented_samples<-subset(inat_clean, user_login %in% overrepresented_users$user_login)
overrepresented_subset<-plyr::ddply(overrepresented_samples,.(user_login),function(x) x[sample(nrow(x),15),])#make the max number of observations per user 15 
normalrepresented_users<-inat_clean %>% group_by(user_login) %>% tally() %>% filter(n<=15)
normalrepresented_samples<-subset(inat_clean, user_login %in% normalrepresented_users$user_login)
inat_clean_corrected<-rbind(overrepresented_subset,normalrepresented_samples)
inat_clean_corrected

inat_clean_tally<-inat_clean %>% group_by(user_login) %>% tally()
mean(inat_clean_tally$n)
median(inat_clean_tally$n)

inatinfec<-inat_clean %>% filter( field.phytoplasma.infection.=="Yes") #make another dataset of infected points
inatinfec %>% group_by(user_login) %>% tally() #vast majority of users contributed less than 6 infected observations, one person contributed 12
inat_clean %>% group_by(user_login) %>% tally()

inat_clean %>% filter(observed_on>'2020-01-01') #23405
23405/28390
```

But maybe also try having size of bubbles indicating number of infected https://stackoverflow.com/questions/60021868/producing-heat-map-over-geo-locations-in-r 

TO DO:
-Turn lat long into equal area projection (eg. 1km by 1km squares), so we have gridded cells. Look at proportion infected to total observations in each cell.
-occupancy model for space: different number of samples per grid, accounting for them
-see what predicts infection?? worldclim data or crop cover? human population density? trillium density? Vector density? ****

## Extracting coordinate data
```{r}
dir.create(path = "data") 
dir.create(path = "output")
bioclim_data_source <- worldclim_global(var = "bio", res = 0.5, path = "data/")
#Determine geographi cextent of our data 
max_lat<-ceiling(max(inat_clean$latitude)) 
min_lat<-floor(min(inat_clean$latitude)) 
max_lon<-ceiling(max(inat_clean$longitude)) 
min_lon<-floor(min(inat_clean$longitude)) 
#Store boundaries in a single extent object 
geographic_extent<-ext(x=c(min_lon,max_lon,min_lat,max_lat))

# Make an extent that is 25% larger 
sample_extent <- geographic_extent * 1.25 
# Crop bioclim data to desired extent 
bioclim_data <- crop(x = bioclim_data_source, y = sample_extent) 
# Plot the first of the bioclim variables to check on cropping 
plot(bioclim_data[[1]])
plot(bioclim_data[[12]])

bioclim_extract <- terra::extract(x = bioclim_data, y = as.matrix(inat_clean[, c("longitude", "latitude")])) # No need for an ID column
# Add the point and climate datasets together 
points_climate <- cbind(inat_clean, bioclim_extract)

crop_data<-cropland(source="WorldCover",  path = "data/")
crop_data<- crop(x = crop_data, y = sample_extent) 
plot(crop_data[[1]])
cropdata_extract<- terra::extract(x = crop_data, y = as.matrix(inat_clean[, c("longitude", "latitude")]))
points <- cbind(points_climate, cropdata_extract)

stonefruit<-crop_monfreda(crop="stonefruitnes", var="area_ha",  path = "data/") #download failed?

footprint<-footprint(year=2009, path = "data/")
footprint_crop<- crop(x = footprint, y = sample_extent) 
plot(footprint_crop[[1]])
footprint_extract<- terra::extract(x = footprint_crop, y = as.matrix(inat_clean[, c("longitude", "latitude")]))
points<-cbind(points, footprint_extract)

bioclim_data
footprint_crop
crop_data
geo_data<-c(bioclim_data, footprint_crop, crop_data)
```

##Density map
```{r}
dat_sf = st_as_sf(inat_clean, coords = c("longitude", "latitude")) %>% 
  st_set_crs("EPSG:4326")
# Reproject to equal area projection
dat_ea = st_transform(dat_sf, crs = "ESRI:102003")
# Make a grid of polygons that encompass the observation points
# Choosing an arbitrary 5 km grid size
area_fishnet_grid = st_make_grid(dat_ea, c(5e4, 5e4), what = "polygons", square = FALSE)

# Convert to sf and add grid ID
fishnet_grid_sf = st_sf(area_fishnet_grid) %>%
  # add grid ID
  mutate(grid_id = 1:length(lengths(area_fishnet_grid)))

# Quick look at the grid
mapview(fishnet_grid_sf)

# Assign observations to grid cells
grid_assignments = st_intersection(fishnet_grid_sf, dat_ea)

# Calculate frequencies per grid cell
freq_per_poly = grid_assignments %>% 
  group_by(grid_id) %>% 
  summarize(n = n(), n_infected = sum(infected == "Yes")) %>% 
  mutate(freq_infected = n_infected/n)

# Rejoin these calculated frequencies to polygons for plotting
freqs_with_polys = st_join(fishnet_grid_sf, freq_per_poly) %>% 
  filter(!is.na(n))

world = ne_countries(scale = "large", returnclass = "sf", continent = "north america")
lakes = ne_download(scale = "medium", type = "lakes", category = "physical")

tril_b<-ggplot() +
  geom_sf(data = world, fill = "white") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  geom_sf(data = freqs_with_polys, aes(fill = freq_infected), alpha = 0.8, color = "lightgrey") +
  scale_fill_gradientn(colours = c("white", "#CCFF99", "darkgreen"), values = c(0,0.000001,1), name="Proportion \ninfected") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()

ggplot() +
  geom_sf(data = world, fill = "white") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  geom_sf(data = freqs_with_polys, aes(fill = n_infected), alpha = 0.8, color = "lightgrey") +
  scale_fill_gradientn(colours = c("white", "#CCFF99", "darkgreen"), values = c(0,0.000001,1), name="Infected observations") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()

tril_a<-ggplot() +
  geom_sf(data = world, fill = "white") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  geom_sf(data = freqs_with_polys, aes(fill = n), alpha = 0.8, color = "lightgrey") +
  scale_fill_gradientn(colours = c("white", "lightblue", "navy"), name="Observations") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()

plot_grid(tril_a, tril_b, labels=c("A", "B"), ncol=1) #10 by 7 portrait
trilab<-plot_grid(tril_a, tril_b, labels=c("A", "B"), ncol=1)
```

##Models
what seems to cause infection?
```{r}
as.data.frame(grid_assignments)
as.data.frame(freq_per_poly)

inat_grid<-merge(as.data.frame(grid_assignments), as.data.frame(freq_per_poly), by.x="grid_id", by.y="grid_id")
inat_grid<-inat_grid %>% select(grid_id, id, n, n_infected, freq_infected)

inat_full<-merge(inat_grid, points, by.x="id", by.y="id")
#inat_full<-rename(inat_full, annual_mean_temp = wc2.1_30s_bio_1, mean_diurnal_range=wc2.1_30s_bio_2, isothermality=wc2.1_30s_bio_3, temp_seasonality=wc2.1_30s_bio_4, max_temp_warmestmonth=wc2.1_30s_bio_5, min_temp_coldestmonth=wc2.1_30s_bio_6, temp_annual_range=wc2.1_30s_bio_7, mean_temp_wettestquart=wc2.1_30s_bio_8, mean_temp_driestquart=wc2.1_30s_bio_9, mean_temp_warmestquart=wc2.1_30s_bio_10, mean_temp_coldestquart=wc2.1_30s_bio_11, annual_precip=wc2.1_30s_bio_12, precip_wettestmonth=wc2.1_30s_bio_13, precip_driestmonth=wc2.1_30s_bio_14, precip_seasonality=wc2.1_30s_bio_15, precip_wettestquarter=wc2.1_30s_bio_16, precip_driestquart=wc2.1_30s_bio_17, precip_warmestquart=wc2.1_30s_bio_18, precip_coldestquart=wc2.1_30s_bio_19)

summary(glmer(as.numeric(infected)~n+scale(annual_mean_temp)+scale(max_temp_warmestmonth)+scale(min_temp_coldestmonth)+scale(annual_precip)+scale(precip_wettestmonth)+scale(precip_driestmonth)+scale(cropland)+scale(`wildareas-v3-2009-human-footprint`)+(1|user_login), family=binomial, data=inat_full)) #adding user login as a random effect makes this take a looooong time

summary(glm(as.numeric(infected) ~ wc2.1_30s_bio_1+ wc2.1_30s_bio_2+wc2.1_30s_bio_3+wc2.1_30s_bio_4+wc2.1_30s_bio_5+wc2.1_30s_bio_6+wc2.1_30s_bio_7+wc2.1_30s_bio_8+wc2.1_30s_bio_9+wc2.1_30s_bio_10+wc2.1_30s_bio_11+wc2.1_30s_bio_12+wc2.1_30s_bio_13+wc2.1_30s_bio_14+wc2.1_30s_bio_15+wc2.1_30s_bio_16+wc2.1_30s_bio_17+wc2.1_30s_bio_18+wc2.1_30s_bio_19+cropland+`wildareas-v3-2009-human-footprint`, data = inat_full, family = binomial))
```
SEM?
```{r}
inat_full$infected<-as.numeric(inat_full$infected)
inat_full$n<-as.numeric(inat_full$n)
inat_full_gridcells<-inat_full %>% group_by(grid_id) %>% 
  slice_sample(n = 1, replace = FALSE)
hist(inat_full_gridcells$n)

summary(glm.nb(n~ wc2.1_30s_bio_1+ wc2.1_30s_bio_2+wc2.1_30s_bio_3+wc2.1_30s_bio_4+wc2.1_30s_bio_5+wc2.1_30s_bio_6+wc2.1_30s_bio_7+wc2.1_30s_bio_8+wc2.1_30s_bio_9+wc2.1_30s_bio_10+wc2.1_30s_bio_11+wc2.1_30s_bio_12+wc2.1_30s_bio_13+wc2.1_30s_bio_14+wc2.1_30s_bio_15+wc2.1_30s_bio_16+wc2.1_30s_bio_17+wc2.1_30s_bio_18+wc2.1_30s_bio_19+cropland+`wildareas-v3-2009-human-footprint`, data=inat_full))

set.seed(1)
modelList_trillium <-  psem(
  glm(infected~n+wc2.1_30s_bio_1+ wc2.1_30s_bio_2+wc2.1_30s_bio_3+wc2.1_30s_bio_4+wc2.1_30s_bio_5+wc2.1_30s_bio_6+wc2.1_30s_bio_7+wc2.1_30s_bio_8+wc2.1_30s_bio_9+wc2.1_30s_bio_10+wc2.1_30s_bio_11+wc2.1_30s_bio_12+wc2.1_30s_bio_13+wc2.1_30s_bio_14+wc2.1_30s_bio_15+wc2.1_30s_bio_16+wc2.1_30s_bio_17+wc2.1_30s_bio_18+wc2.1_30s_bio_19+cropland+`wildareas-v3-2009-human-footprint`, family=binomial, data=inat_full),
  glm.nb(n~ wc2.1_30s_bio_1+ wc2.1_30s_bio_2+wc2.1_30s_bio_3+wc2.1_30s_bio_4+wc2.1_30s_bio_5+wc2.1_30s_bio_6+wc2.1_30s_bio_7+wc2.1_30s_bio_8+wc2.1_30s_bio_9+wc2.1_30s_bio_10+wc2.1_30s_bio_11+wc2.1_30s_bio_12+wc2.1_30s_bio_13+wc2.1_30s_bio_14+wc2.1_30s_bio_15+wc2.1_30s_bio_16+wc2.1_30s_bio_17+wc2.1_30s_bio_18+wc2.1_30s_bio_19+cropland+`wildareas-v3-2009-human-footprint`, data=inat_full)
 )
summary(modelList_trillium)
fisherC(modelList_trillium) 
modelList_trillium_coefs<-as.data.frame(coefs(modelList_trillium))
modelList_trillium_coefs<-modelList_trillium_coefs[,c(2,1, 3:8)]
modelList_trillium_coefssig<-modelList_trillium_obs_coefs %>% filter(P.Value<0.05)
modelList_trillium_coefssig
```

Species distribution model?
https://jcoliver.github.io/learn-r/011-species-distribution-models.pdf 
```{r}
inat_1<-merge(inat_grid, points, by.x="id", by.y="id")
inat_1 = subset(inat_1, select = -c(.val,.equ, .zer, .cen, .inst, .dpl, .summary) )
# Create vector indicating fold for making a training dataset 
fold <- folds(x = inat_1, k = 5, by = inat_full$infected) #five folds so that we can test on 20% of the data
testing <- inat_1[fold == 1, ] 
training <- inat_1[fold != 1, ]

# Build a model using training data 
glm_model <- glm(as.numeric(infected) ~ wc2.1_30s_bio_1+ wc2.1_30s_bio_2+wc2.1_30s_bio_3+wc2.1_30s_bio_4+wc2.1_30s_bio_5+wc2.1_30s_bio_6+wc2.1_30s_bio_7+wc2.1_30s_bio_8+wc2.1_30s_bio_9+wc2.1_30s_bio_10+wc2.1_30s_bio_11+wc2.1_30s_bio_12+wc2.1_30s_bio_13+wc2.1_30s_bio_14+wc2.1_30s_bio_15+wc2.1_30s_bio_16+wc2.1_30s_bio_17+wc2.1_30s_bio_18+wc2.1_30s_bio_19+cropland+`wildareas-v3-2009-human-footprint`, data = training, family = binomial)
summary(glm_model)

# Get predicted values from the model 
glm_predict <- predict(geo_data, glm_model, type = "response") #should I incorporate trillium n into this?
# Print predicted values 
plot(glm_predict)
# Use testing data for model evaluation 
glm_eval <- pa_evaluate(p = testing[testing$infected == 1, ], a = testing[testing$infected == 0, ], model = glm_model, type = "response")
# Determine minimum threshold for "presence" 
glm_threshold <- glm_eval@thresholds$max_spec_sens
```
```{r}
world = ne_countries(scale = "large", returnclass = "sf", continent = "north america")
lakes = ne_download(scale = "medium", type = "lakes", category = "physical")

tril_c<-ggplot() +
   geom_spatraster(data = (glm_predict> glm_threshold), aes(fill = lyr1))+#Only plot areas where probability of occurrence is greater than the threshold 
  geom_sf(data = world, fill="NA") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  coord_sf(xlim =c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()+scale_fill_manual(name="Suitability for phytoplasma", values=c("NA","red"), na.value="NA")+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), shape=3, size=1,alpha=0.3)
#the map we have drawn presents a categorical classification of whether a particular point on the landscape will be suitable or not for the species of interest.
```

```{r}
tril_a
tril_b
tril_c
plot_grid(trilab, tril_c, labels=c("","C"), ncol=2)
```


##############################################################################################
##Old Main maps
```{r}
#http://mbontrager.org/blog/2016/08/01/Maps-in-R
infectedpoints <- filter(inat_clean_corrected, field.phytoplasma.infection.=="Yes")
normalpoints <- filter(inat_clean_corrected, field.phytoplasma.infection.=="")
infcolour <- brewer.pal(6, "Paired")[5]
normcolour <- brewer.pal(6, "Paired")[3]

usa <- map_data("state")
canada <- map_data("worldHires","Canada")

NAmap <- ggplot() + geom_polygon(data = usa, 
                                 aes(x=long, y = lat, group = group), 
                                 fill = "white", 
                                 color="black") +
    geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
                 fill = "white", color="black") +
    coord_fixed(xlim = c(-95, -70),  ylim = c(33, 50), ratio = 1.2)
NAmap

range<-NAmap+ geom_point(data=inat_clean_corrected, aes(x=longitude, y=latitude), fill="hotpink",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "*Trillium grandiflorum* observations") +theme(title = ggtext::element_markdown())
infections<-NAmap+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), fill="yellowgreen",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "Infected *T. grandiflorum* observations") +theme(title = ggtext::element_markdown())
rangemap<-plot_grid(range,infections, labels=c("A", "B"))
ggsave2("phytoplasma_rangemap.png", rangemap, dpi=600, width=10, height=8)
```

tmap version
https://edzer.github.io/sdsr_exercises/08.html
```{r}
us_states_map = tm_shape(us_states, crs = "EPSG:9822") + 
  tm_polygons() + 
  tm_layout(frame = FALSE)


ne = ne_countries(returnclass = "sf")
ne |> filter(admin == "Canada") -> canada
ca_eqc = st_transform(canada,"EPSG:9822")
ne |> filter(admin == "United States of America") -> us
us_eqc = st_transform(us,"EPSG:9822")
plot(us_eqc)

nz_elev = rast(system.file("raster/nz_elev.tif", package = "spDataLarge"))
```

sf version with ggplot
```{r}
st_layers("gadm41_CAN.gpkg")
canada <- read_sf("gadm41_CAN.gpkg", layer = "ADM_ADM_1") #https://inbo.github.io/tutorials/tutorials/spatial_standards_vector/

#canada <- read_sf("gadm41_CAN_1.shp")
#gadm41_CAN_1.shp
st_layers("gadm41_USA.gpkg")
us <- read_sf("gadm41_USA.gpkg", layer = "ADM_ADM_1")
plot(canada)
plot(us)
#b = st_buffer(canada[1,1], units::set_units(10, km)) #|> st_cast("LINESTRING")

#https://www.naturalearthdata.com/downloads/10m-physical-vectors/10m-lakes/
# Create temp files
temp2 <- tempfile()
# Download the zip file and save unzipped
# Unzip the contents of the temp and save unzipped content in 'temp2'
unzip(zipfile = "ne_10m_lakes.zip", exdir = temp2)
# Read the shapefile. Alternatively make an assignment, such as f<-sf::read_sf(your_SHP_file)
lakes<-sf::read_sf(temp2)

rangev2<-ggplot() + geom_sf(data = canada, fill="white") + geom_sf(data = us, fill="white")+geom_sf(data=lakes)+coord_sf(xlim = c(-95, -70),  ylim = c(33, 50))+ geom_point(data=inat_clean_corrected, aes(x=longitude, y=latitude), fill="hotpink",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "*Trillium grandiflorum* observations") +theme(title = ggtext::element_markdown()) 
#THIS MADE ME NOTICE A FEW OBSERVATIONS IN LAKE ONTARIO- FIND AND REMOVE!
#could use county map and colour by number of observations within a given county?

infectionsv2<-ggplot() + geom_sf(data = canada, fill="white") + geom_sf(data = us, fill="white")+geom_sf(data=lakes)+coord_sf(xlim = c(-95, -70),  ylim = c(33, 50))+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), fill="limegreen",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "Infected *T. grandiflorum* observations") +theme(title = ggtext::element_markdown())
rangemapv2<-plot_grid(rangev2,infectionsv2, labels=c("A", "B"))
ggsave2("phytoplasma_rangemapV2.png", rangemapv2, dpi=600, width=10, height=8)
```
```{r}
#https://www.naturalearthdata.com/downloads/10m-cultural-vectors/parks-and-protected-lands/
#https://www.naturalearthdata.com/downloads/10m-cultural-vectors/10m-urban-area/ 
```

Equal area projections
https://r.geocompx.org/reproj-geo-data.html
https://bbest.github.io/R-adv-spatial-lessons/tidy.html#challenge-project-recalculate-area
https://epsg.io/9822-method
https://www.andrewheiss.com/blog/2023/07/28/gradient-map-fills-r-sf/ !!!
```{r}
library(geosphere)
st_crs(us) # get CRS
st_crs(canada)
st_crs(lakes)

crs_usa = '+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs'
usa_geo = us %>%
  st_transform(st_crs("EPSG:9822"))
st_crs(us)
st_crs(usa_geo)
canada_geo = canada %>%
  st_transform(st_crs("EPSG:9822"))
lakes_geo = lakes %>%
  st_transform(st_crs("EPSG:9822"))
basemap<-canada_geo+usa_geo+lakes_geo

ggplot() + geom_sf(data = canada, fill="white") + geom_sf(data = us, fill="white")+geom_sf(data=lakes)+coord_sf(xlim = c(-95, -70),  ylim = c(33, 50))+ geom_point(data=inat_clean_corrected, aes(x=longitude, y=latitude), fill="hotpink",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "*Trillium grandiflorum* observations") +theme(title = ggtext::element_markdown()) 
#THIS MADE ME NOTICE A FEW OBSERVATIONS IN LAKE ONTARIO- FIND AND REMOVE!
#could use county map and colour by number of observations within a given county?
ggplot() + geom_sf(data = canada_geo, fill="white") + geom_sf(data = usa_geo, fill="white")+geom_sf(data=lakes_geo)+coord_sf(xlim = c(-95, -70),  ylim = c(33, 50), crs = "EPSG:9822")+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), fill="limegreen",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "Infected *T. grandiflorum* observations") +theme(title = ggtext::element_markdown())

infection_coords_geo <- st_as_sf(x = inatinfec,                         
           coords = c("longitude", "latitude"),
           crs = "EPSG:9822")
observation_coords_geo <- st_as_sf(x = inat_clean_corrected,                         
           coords = c("longitude", "latitude"),
           crs = "EPSG:9822")
```
density
```{r}
library(spatstat)
infection_ppp <- as.ppp(infection_coords_geo, W = as.owin(canada_geo))
# Check to see if it worked
plot(infection_ppp)
infection_ppp_stars <- stars::st_as_stars(density(infection_ppp, dimyx = 300))
```



