---
title: "trillium_phytoplasma"
author: "Julia Boyle"
date: "2023-05-27"
output: html_document
---
# iNaturalist data
```{r}
#install.packages('geodata')
library(maps)
library(mapdata)
library(ggmap)
library(RColorBrewer)
library(raster)
library(scales)
library(CoordinateCleaner)
library(rnaturalearth)
library(rnaturalearthdata)
library(geodata)
library(sf)
library(plyr)
library(spData)
library(tidyterra)
library(terra)
library(tmap) 
library(predicts)
library(tidyverse)
library(cowplot)
```

```{r}
#inat<-read.csv("iNat_alltrilliums_2023_07_05.csv") #only up until 2023
inat<-read.csv("iNat_alltrilliums_2025_03_14_539340.csv") #version with 2024 data
head(inat)

inat_cleaned = inat %>% 
  # Adding positional accuracy filter to see if that helps deal with observations in lakes
  filter(!is.na(latitude), !is.na(longitude), !is.na(public_positional_accuracy)) %>% 
  filter(public_positional_accuracy < 10000) %>% 
  dplyr::mutate(infected = `field.phytoplasma.infection.`) %>% 
  mutate(infected = if_else(is.na(infected), "No", infected)) %>% 
  # Drop european and puerto rico observations
  filter(latitude > 20, longitude < -65) %>% 
  # Group by user and subset to 15 observations per user
  group_by(user_login) %>% 
  slice_sample(n = 15, replace = FALSE)

inat_clean <- clean_coordinates(x = inat_cleaned, 
                           lon = "longitude", 
                           lat = "latitude",
                           species = "scientific_name",
                           tests = c("centroids",
                                    "equal", "zeros","institutions", "duplicates")) 
inat_clean=inat_clean %>% filter(.summary==TRUE) #only include observations that have no suspect issues
inat_clean=inat_clean %>% mutate(infected=ifelse(grepl("Yes", infected), "1", "0"))
inat_clean$infected

inat%>% filter( field.phytoplasma.infection.=="Yes") #845
inat_cleaned%>% filter( field.phytoplasma.infection.=="Yes") #550
inat_clean%>% filter( field.phytoplasma.infection.=="Yes") #526 after filtering out stuff with clean_coordinates
```
## Extracting coordinate data
Run this everytime you load R because spatrasters do not save with R enviro
```{r}
dir.create(path = "data") 
dir.create(path = "output")
bioclim_data_source <- worldclim_global(var = "bio", res = 0.5, path = "data/")
#Determine geographi cextent of our data 
max_lat<-ceiling(max(inat_clean$latitude)) 
min_lat<-floor(min(inat_clean$latitude)) 
max_lon<-ceiling(max(inat_clean$longitude)) 
min_lon<-floor(min(inat_clean$longitude)) 
#Store boundaries in a single extent object 
geographic_extent<-ext(x=c(min_lon,max_lon,min_lat,max_lat))

# Make an extent that is 25% larger 
sample_extent <- geographic_extent * 1.25 
# Crop bioclim data to desired extent 
bioclim_data <- crop(x = bioclim_data_source, y = sample_extent) 
# Plot the first of the bioclim variables to check on cropping 
plot(bioclim_data[[1]])
plot(bioclim_data[[12]])

bioclim_extract <- terra::extract(x = bioclim_data, y = as.matrix(inat_clean[, c("longitude", "latitude")])) # No need for an ID column
# I GET 41 OBSERVATIONS WITH NAS FROM BIOCLIM DATA- WHY???
# Add the point and climate datasets together 
points_climate <- cbind(inat_clean, bioclim_extract)

crop_data<-cropland(source="WorldCover",  path = "data/")
crop_data<- crop(x = crop_data, y = sample_extent) 
plot(crop_data[[1]])
cropdata_extract<- terra::extract(x = crop_data, y = as.matrix(inat_clean[, c("longitude", "latitude")]))
points <- cbind(points_climate, cropdata_extract)

footprint<-footprint(year=2009, path = "data/")
footprint_crop<- crop(x = footprint, y = sample_extent) 
plot(footprint_crop[[1]])
footprint_extract<- terra::extract(x = footprint_crop, y = as.matrix(inat_clean[, c("longitude", "latitude")]))
footprint_extract<-rename(footprint_extract, footprint = `wildareas-v3-2009-human-footprint`
)
points<-cbind(points, footprint_extract)

bioclim_data
footprint_crop
crop_data
geo_data<-c(bioclim_data, footprint_crop, crop_data)
geo_data_test<-geo_data
```

##Density map
```{r}
dat_sf = st_as_sf(inat_clean, coords = c("longitude", "latitude")) %>% 
  st_set_crs("EPSG:4326")
# Reproject to equal area projection
dat_ea = st_transform(dat_sf, crs = "ESRI:102003")
# Make a grid of polygons that encompass the observation points
# Choosing an arbitrary 5 km grid size
area_fishnet_grid = st_make_grid(dat_ea, c(5e4, 5e4), what = "polygons", square = FALSE)

# Convert to sf and add grid ID
fishnet_grid_sf = st_sf(area_fishnet_grid) %>%
  # add grid ID
  mutate(grid_id = 1:length(lengths(area_fishnet_grid)))
# Assign observations to grid cells
grid_assignments = st_intersection(fishnet_grid_sf, dat_ea)

# Calculate frequencies per grid cell
freq_per_poly = grid_assignments %>% 
  group_by(grid_id) %>% 
  summarize(n = n(), n_infected = sum(infected == "1")) %>% 
  mutate(freq_infected = n_infected/n)

# Rejoin these calculated frequencies to polygons for plotting
freqs_with_polys = st_join(fishnet_grid_sf, freq_per_poly) %>% 
  filter(!is.na(n))

world = ne_countries(scale = "large", returnclass = "sf", continent = "north america")
lakes = ne_download(scale = "medium", type = "lakes", category = "physical")

tril_b<-ggplot() +
  geom_sf(data = world, fill = "white") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  geom_sf(data = freqs_with_polys, aes(fill = freq_infected), alpha = 0.8, color = "lightgrey") +
  scale_fill_gradientn(colours = c("white", "#CCFF99", "darkgreen"), values = c(0,0.000001,1), name="Proportion \ninfected") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()

tril_a<-ggplot() +
  geom_sf(data = world, fill = "white") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  geom_sf(data = freqs_with_polys, aes(fill = n), alpha = 0.8, color = "lightgrey") +
  scale_fill_gradientn(colours = c("white", "lightblue", "navy"), name="Observations") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()

plot_grid(tril_a, tril_b, labels=c("A", "B"), ncol=1) #10 by 7 portrait
trilab<-plot_grid(tril_a, tril_b, labels=c("A", "B"), ncol=1)
```
##Moran’s I analysis
https://mgimond.github.io/simple_moransI_example/
```{r}
library(sf)
library(spdep)
library(tmap)
nb <- poly2nb(freqs_with_polys, queen=TRUE)
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
inc.lag <- lag.listw(lw, freqs_with_polys$n_infected)
plot(inc.lag ~ freqs_with_polys$n_infected, pch=16, asp=1)
M1 <- lm(inc.lag ~ freqs_with_polys$n_infected)
abline(M1, col="blue")
coef(M1)[2] #0.4216945
#this slope is Morans I: because this is positive, it means polygons with high frequency of infection are more likely to have neighbouring polygons with high frequency of infection
#a positive Moran’s I value suggests clustering whereas a negative Moran’s I value suggests dispersion
inc.lag.freq <- lag.listw(lw, freqs_with_polys$freq_infected)
plot(inc.lag.freq ~ freqs_with_polys$freq_infected, pch=16, asp=1)
M1.freq <- lm(inc.lag.freq ~ freqs_with_polys$freq_infected)
abline(M1.freq, col="blue")
coef(M1.freq)[2] #0.1072854

moran(freqs_with_polys$n_infected, lw, length(nb), Szero(lw))[1] #0.4302413, infections are pretty clustered in space
moran(freqs_with_polys$freq_infected, lw, length(nb), Szero(lw))[1] #0.1082928
moran(freqs_with_polys$n, lw, length(nb), Szero(lw))[1] #0.3096669

MC.ninfec<- moran.mc(freqs_with_polys$n_infected, lw, nsim=999, alternative="greater")
MC.ninfec
plot(MC.ninfec)
MC.freq<- moran.mc(freqs_with_polys$freq_infected, lw, nsim=999, alternative="greater")
MC.freq
plot(MC.freq)
MC.n<- moran.mc(freqs_with_polys$n, lw, nsim=999, alternative="greater")
MC.n
plot(MC.n)
```

##Models
what seems to cause infection?
```{r}
as.data.frame(grid_assignments)
as.data.frame(freq_per_poly)

inat_grid<-merge(as.data.frame(grid_assignments), as.data.frame(freq_per_poly), by.x="grid_id", by.y="grid_id")
inat_grid<-inat_grid %>% dplyr::select(grid_id, id, n, n_infected, freq_infected)

inat_full<-merge(inat_grid, points, by.x="id", by.y="id")
inat_full$infected<-as.numeric(inat_full$infected)
#inat_full$footprint<-inat_full$`wildareas-v3-2009-human-footprint`
inat_full$Year<-substring(inat_full$observed_on,1,4)
str(inat_full)
```

###GAM

Run models
```{r}
library(mgcv)
#https://m-clark.github.io/generalized-additive-models/appendix.html
#https://r.qcbs.ca/workshop08/book-en/gam-model-checking.html
#https://noamross.github.io/gams-in-r-course/     <-----THIS ONE IS GREAT!!!
# +s(latitude,longitude,bs="sos",m=2,k=200) #Spherical version
#+s(latitude,longitude,bs="gp",m=2,k=200) #gaussian version
gam_tril_allsmooth<-gam(infected ~ s(n, k=50) +s(wc2.1_30s_bio_1) +s(wc2.1_30s_bio_4) +s(wc2.1_30s_bio_5) +s(wc2.1_30s_bio_6) +s(wc2.1_30s_bio_12) +s(wc2.1_30s_bio_13) +s(wc2.1_30s_bio_14) +s(wc2.1_30s_bio_15) +s(cropland) +s(footprint) +s(latitude,longitude,bs="gp",m=2,k=200),family="binomial",method = "REML",data=inat_full)

#model 1
summary(gam_tril_allsmooth)
k.check(gam_tril_allsmooth) #if edf is too close to k, gives sig value. we want k-index near 1 and nonsig p. k controls wiggliness
set.seed(1)
gam.check(gam_tril_allsmooth)
plot(gam_tril_allsmooth, residuals = TRUE, pch = 1, all.terms = TRUE, shade = TRUE, shift = coef(gam_tril)[1])
concurvity(gam_tril_allsmooth, full = TRUE) #if worst case has values over 0.8 then you have problem similar to colinearity but with parabola-ness. Use full=F to find out the offenders
concurvity(gam_tril_allsmooth, full = F)
rpred <- predict(geo_data, gam_tril_allsmooth)#but maybe add one layer with tril n. eg. rstack <- stack(rx1, rx2, rzeros)/ names(rstack) <- c("x1", "x2", "sitef")
vis.gam(gam_tril_allsmooth, view = c("latitude", "longitude"), plot.type = "contour", too.far = 0.05)#means it won't extrapolate more than 5% into untested combinations

#model 2
set.seed(1)
gam_vifstep<-gam(infected ~ s(n, k=50)+ s(wc2.1_30s_bio_2)+s(wc2.1_30s_bio_7)+s(wc2.1_30s_bio_8)+s(wc2.1_30s_bio_9)+s(wc2.1_30s_bio_10)+s(wc2.1_30s_bio_15)+s(wc2.1_30s_bio_18, k=30)+s(cropland)+s(footprint)+s(latitude,longitude,bs="gp",m=2,k=200),family="binomial",method = "REML",data=inat_full)
summary(gam_vifstep)
gam.check(gam_vifstep)
vis.gam(gam_vifstep, view = c("longitude", "latitude"), plot.type = "contour", too.far = 0.05, color="topo")#means it won't extrapolate more than 5% into untested combinations
plot(gam_vifstep, residuals = TRUE, pch = 1, all.terms = TRUE, shade = TRUE, shift = coef(gam_tril)[1])
```

Species distribution model?
https://jcoliver.github.io/learn-r/011-species-distribution-models.pdf 
```{r}
inat_1<-merge(inat_grid, points, by.x="id", by.y="id")
inat_1 = subset(inat_1, select = -c(.val,.equ, .zer, .cen, .inst, .dpl, .summary) )
# Create vector indicating fold for making a training dataset 
fold <- folds(x = inat_1, k = 5, by = inat_full$infected) #five folds so that we can test on 20% of the data
testing <- inat_1[fold == 1, ] 
training <- inat_1[fold != 1, ]
training$footprint<-as.numeric(training$footprint)
training$infected<-as.numeric(training$infected)
str(training)
# Build a model using training data 
gam_training<-gam(infected ~ s(n, k=50)+ s(wc2.1_30s_bio_2)+s(wc2.1_30s_bio_7)+s(wc2.1_30s_bio_8)+s(wc2.1_30s_bio_9)+s(wc2.1_30s_bio_10)+s(wc2.1_30s_bio_15)+s(wc2.1_30s_bio_18, k=30)+s(cropland)+s(footprint)+s(latitude,longitude,bs="gp",m=2,k=200),family="binomial",method = "REML",data=training)
summary(gam_training)

which(is.na(training$footprint))
names(geo_data_test)
# Get predicted values from the model 
gam_predict <- predict(geo_data_test, gam_training, type = "response") #geo_data but with trillium number and lat and long
# Print predicted values 
plot(gam_predict)
# Use testing data for model evaluation 
gam_eval <- pa_evaluate(p = testing[testing$infected == 1, ], a = testing[testing$infected == 0, ], model = gam_training, type = "response")
# Determine minimum threshold for "presence" 
gam_threshold <- gam_eval@thresholds$max_spec_sens
```
```{r}
world = ne_countries(scale = "large", returnclass = "sf", continent = "north america")
lakes = ne_download(scale = "medium", type = "lakes", category = "physical")

tril_c<-ggplot() +
   geom_spatraster(data = (gam_predict> gam_threshold), aes(fill = lyr1))+#Only plot areas where probability of occurrence is greater than the threshold 
  geom_sf(data = world, fill="NA") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  coord_sf(xlim =c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()+scale_fill_manual(name="Suitability for phytoplasma", values=c("NA","red"), na.value="NA")+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), shape=3, size=1,alpha=0.3)
#the map we have drawn presents a categorical classification of whether a particular point on the landscape will be suitable or not for the species of interest.
```


### INLA
https://ourcodingclub.github.io/tutorials/inla/
https://ourcodingclub.github.io/tutorials/spatial-modelling-inla/
```{r}
#install.packages("devtools")
#if(!require(ggregplot)) devtools::install_github("gfalbery/ggregplot") # Installing Greg's package for plotting functions!
#install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA); library(ggregplot); library(RColorBrewer)
library(INLA)
library(INLAspacetime)
library(usdm)
head(inat_full)

##########################################################################
which(is.na(inat_full$wc2.1_30s_bio_1)) #41 uninfected trillium near quebec and montreal don't have associated bioclim data for some reason. we need to exclude them for INLA
inat_full[which(is.na(inat_full$wc2.1_30s_bio_1)),]
inat_full_complete = inat_full[complete.cases(inat_full[53:73]),]
##########################################################################

phen <- c("longitude", "latitude") # Base columns with spatial information we'll need

resp <- "infected" # Response variable

covar <- c("n","wc2.1_30s_bio_1",  "wc2.1_30s_bio_2","wc2.1_30s_bio_3","wc2.1_30s_bio_4","wc2.1_30s_bio_5","wc2.1_30s_bio_6","wc2.1_30s_bio_7","wc2.1_30s_bio_8","wc2.1_30s_bio_9","wc2.1_30s_bio_10","wc2.1_30s_bio_11","wc2.1_30s_bio_12","wc2.1_30s_bio_13","wc2.1_30s_bio_14","wc2.1_30s_bio_15","wc2.1_30s_bio_16","wc2.1_30s_bio_17","wc2.1_30s_bio_18","wc2.1_30s_bio_19","cropland", "footprint")

#Selecting final variables for the model
ggregplot::INLAModelSel(resp, covar, Family="binomial", Data=inat_full_complete) #removes variables that improve DIC by at least 2
  #[1] "Losing wc2.1_30s_bio_18"[1] "Run 2"[1] "Losing wc2.1_30s_bio_1"[1] "Run 3"[1] "Losing wc2.1_30s_bio_13"[1] "Run 4" "Losing wc2.1_30s_bio_16"[1] "Run 5[1]   "Losing wc2.1_30s_bio_17"[1] "Run 6"[1] "Losing n"[1] "Run 7"[1] "Losing wc2.1_30s_bio_5"[1] "Run 8"[1] "Losing footprint"[1] "Run 9"[1] "Losing   wc2.1_30s_bio_11"[1] "Run 10"[1] "Losing wc2.1_30s_bio_15"[1] "Run 11" 
  #it removed n and footprint, but I want to keep them in the model due to a priori reasons

vifcor(inat_full_complete %>% dplyr::select(n,wc2.1_30s_bio_2,wc2.1_30s_bio_3,wc2.1_30s_bio_4,wc2.1_30s_bio_6,wc2.1_30s_bio_7,wc2.1_30s_bio_8,wc2.1_30s_bio_9,wc2.1_30s_bio_10,wc2.1_30s_bio_12,wc2.1_30s_bio_14,wc2.1_30s_bio_19,cropland, footprint)) #remove vars 19, 7, 4, 12 due to multicollinearity #except var 3 still has REALLY HIGH VIF! Let's see what happens when we remove it
vifcor(inat_full_complete %>% dplyr::select(n,wc2.1_30s_bio_2,wc2.1_30s_bio_6,wc2.1_30s_bio_8,wc2.1_30s_bio_9,wc2.1_30s_bio_10,wc2.1_30s_bio_14,cropland, footprint))#After removing var 3, all VIFs are under 5, which is great

Finalcovar<- c("n","wc2.1_30s_bio_2","wc2.1_30s_bio_6","wc2.1_30s_bio_8","wc2.1_30s_bio_9","wc2.1_30s_bio_10","wc2.1_30s_bio_14","cropland", "footprint")

#incorporating space
Locations = cbind(inat_full_complete$longitude, inat_full_complete$latitude) # using the sampling locations
Mesh <- INLA::inla.mesh.2d(Locations, max.edge = c(10, 500), cutoff=1) #smaller triangles- use smaller triangles for publication
plot(Mesh)
points(Locations, col = "red", pch = 2)
# Making the A matrix
HostsA <- INLA::inla.spde.make.A(Mesh, loc = Locations) # Making A matrix, which essentially translates spatial locations on the mesh into vectors in the model.
Range_prior<-inat_full_complete[, 28:29]%>% slice(1:n(), 100)%>% dist%>% c%>% max%>% divide_by(2) #14.72662. This is the null range prior. This method of determining the prior is found from code in https://doi.org/10.1098/rspb.2020.2655
Hosts.spde = INLA::inla.spde2.pcmatern(mesh = Mesh, prior.range = c(Range_prior, 0.5), prior.sigma = c(.5, .5)) 
w.Host <- INLA::inla.spde.make.index('w', n.spde = Hosts.spde$n.spde) # making the w

# Making the model matrix #### 
X0 <- model.matrix(as.formula(paste0("~ -1 +", paste(Finalcovar, collapse = "+"))), data = inat_full_complete) # make the model matrix using the final model selection formula without a response variable.
X <- as.data.frame(X0)
# Making the stack ####
N <- nrow(inat_full_complete)
StackHost <- INLA::inla.stack(
  data = list(y = inat_full_complete[,resp]), # specify the response variable
  A = list(1,1,HostsA), # Vector of Multiplication factors for random and fixed effects  #put as many 1s as there are fixed and random effects in the effects() below            
  effects = list(
    Intercept = rep(1, N), # specify the manual intercept!
    X = X, # attach the model matrix
    # insert vectors of any random effects (none here)
    w = w.Host)) # attach the w 
#Running the model
f1 <- as.formula(paste0("y ~ -1 + Intercept + ", paste0(colnames(X), collapse = " + "))) #version only fixed effects
f3 <- as.formula(paste0("y ~ -1 + Intercept + ", paste0(colnames(X), collapse = " + "), " + f(w, model = Hosts.spde)")) #version with space SPDE random effect
IM1 <- INLA::inla(f1, # Base model (no random effects)
            family = "binomial",
            data = inla.stack.data(StackHost),
            control.compute = list(dic = TRUE),
            control.predictor = list(A = INLA::inla.stack.A(StackHost)))
IM3 <- INLA::inla(f3, # SPDE random effect 
            family = "binomial",
            data = inla.stack.data(StackHost),
            control.compute = list(dic = TRUE),
            control.predictor = list(A = INLA::inla.stack.A(StackHost)))
SpatialHostList <- list(IM1, IM3)

ggregplot::ggField(IM3, Mesh, Groups = 1) +
  scale_fill_brewer(palette = "Blues")#Plotting the spatial field 
sapply(SpatialHostList, function(f) f$dic$dic) #does space make it better?
ggregplot::INLADICFig(SpatialHostList, ModelNames = c("Base", "SPDE")) #if DIC is smaller by >2, then suggests autocorrelation was important to include
dput(summary(IM3), file="INLAsummary_IM3.txt", control="all")
summary(IM3)#There are no P values in INLA. Importance or significance of variables can be deduced by examining the overlap of their 2.5% and 97.5% posterior estimates with zero.
ggregplot::Efxplot(SpatialHostList)
ggregplot::Efxplot(IM3)+ggtitle("INLA spatial model")
saveRDS(SpatialHostList, paste0("output/","SpatialHostList",".rds"))
saveRDS(IM3, paste0("output/","IM3",".rds"))
```
https://ourcodingclub.github.io/tutorials/spatial-modelling-inla/
Making predictions
"To produce predictions using INLA, we need to generate a dataset (with attached coordinates on the locations we wish to predict to) and attach a series of missing observation to it (coded as NA in R). When the missing observations are in the response variable, INLA automatically computes the predictive distribution of the corresponding linear predictor and fitted values.
Using INLA syntax is possible to generate model predictions by fitting a stack where the response variable is set as NAs, and then join this stack with the estimation stack (which is similar to what we have used so far). Then we can extract the values of the predicted response variable and use the inla.mesh.projector() function to project these values on the mesh vertices (like we have been doing when plotting the GRF earlier on)."
```{r}
# The first step is to load the prediction raster file (this one is a ASCII file).
require(raster)
require(rgdal)
#put in my geo_data_test file here
GS_Matrix <- matrix(crop_data) #To start, we transform the raster values for the amount of green space (GS ratio) into a matrix and then reallocate the coordinates to a matrix of ncol X nrow cells (numbers of columns and rows).
GS_Pred<-raster(GS_Matrix)
str(GS_Matrix)

y.res <- GS_Pred@nrows
x.res <- GS_Pred@ncols
#Next, we need to create a grid of ncol X nrow cells containing the coordinates of the points where we wish to project our model predictions.
Seq.X.grid <- seq(from = GS_Pred@extent@xmin,
                  to = GS_Pred@extent@xmax,
                  length = x.res)

Seq.Y.grid <- seq(from = GS_Pred@extent@ymin,
                  to = GS_Pred@extent@ymax,
                  length = y.res)

pred.grid <- as.matrix(expand.grid(x = Seq.X.grid,
                                   y = Seq.Y.grid))

str(pred.grid)
#Now that we the grid with the coordinates of each cell centroid we can procede to make the mesh SPDE and spatial index as usual.
MeshPred <- inla.mesh.2d(Locations, max.edge = c(10, 500), cutoff=1)
plot(MeshPred)

spde.pred <- inla.spde2.matern(mesh = MeshPred,
                               alpha = 2)

s.index.p <- inla.spde.make.index(name = "sp.field.pred",
                                  n.spde = spde.pred$n.spde)
#Since the points where we want to project our predictions are different from the datapoints, we need two different projector matrices. The first one is the standard one we have used so far (A_est), while the second does not contain point locations since we will project the model results directly on the mesh vertices. Similarly, we will need two stacks, one for estimations and one for predictions, joined using the inla.stack() function to form a joined stack
A_est <- inla.spde.make.A(mesh = MeshPred,
                          loc = Locations)

A_pred <- inla.spde.make.A(mesh = MeshPred)

StackEst <- INLA::inla.stack(
  data = list(y = inat_full_complete[,resp]), # specify the response variable
  A = list(1,1,HostsA), # Vector of Multiplication factors for random and fixed effects  #put as many 1s as there are fixed and random effects in the effects() below            
  effects = list(
    Intercept = rep(1, N), # specify the manual intercept!
    X = X, # attach the model matrix
    # insert vectors of any random effects (none here)
    w = w.Host),tag = "Est")

stackPred <- inla.stack(data = list(y = NA),  # NAs in the response variable  
                        A = list(1,1,A_pred),
                        effects = list( Intercept = rep(1, N), X = X,  w = w.Host), tag = "Pred")

StackJoin <- inla.stack(StackEst, stackPred)
#Then we can specify the formula and run the model as usual (using the joint stack).
Mod_Pred<-INLA::inla(f3, # SPDE random effect 
            family = "binomial",
            data = inla.stack.data(StackJoin),
            control.compute = list(dic = TRUE),
            control.predictor = list(A = INLA::inla.stack.A(StackJoin),  compute = T))
#We need to extract the index of the data from the prediction part of the stack (using the tag “Pred” we assigned to the stack) and use it to select the relevant posterior mean and sd for the predicted response variable. Then we use the inla.mesh.projector() function to calculate the projection from the Mesh to the grid we created (pred.grid).
index.pred <- inla.stack.index(StackJoin, "Pred")$data

post.mean.pred <- Mod_Pred$summary.linear.predictor[index.pred, "mean"]
post.sd.pred <- Mod_Pred$summary.linear.predictor[index.pred, "sd"]

proj.grid <- inla.mesh.projector(MeshPred,
                                 xlim = range(pred.grid[,1]),
                                 ylim = range(pred.grid[,2]),
                                 dims = c(x.res, y.res))
#Finally, we project the values we extracted from the model on the lattice we have created and transform the projected predictions to a raster object and plot them (we do this for both the mean and standard deviation).
post.mean.pred.grid <- inla.mesh.project(proj.grid, post.mean.pred)
post.sd.pred.grid <- inla.mesh.project(proj.grid, post.sd.pred)

predmean <- t(post.mean.pred.grid)
predmean2 <- predmean[rev(1:length(predmean[,1])),]
predmean_ras <- raster(predmean2,
                      xmn = range(projgrid$x)[1], xmx = range(projgrid$x)[2],
                      ymn = range(projgrid$y)[1], ymx = range(projgrid$y)[2],
                      crs = CRS("ESRI:102003"))

predsd <- t(post.sd.pred.grid)
predsd2 <- predsd[rev(1:length(predsd[,1])),]
predsd_ras <- raster(predsd2,
                       xmn = range(projgrid$x)[1], xmx = range(projgrid$x)[2],
                       ymn = range(projgrid$y)[1], ymx = range(projgrid$y)[2],
                       crs = CRS("ESRI:102003"))

# plot the model predictions for mean
par(mfrow = c(1,1), mar = c(2,2, 1,1))
plot(predmean_ras, asp = 1, col = my.palette.post)
points(Fox_Point, pch = 16, cex = 0.5)
#plot(Scot_Shape_BNG, add = T) #add map using shapefile of NA

# plot the model predictions for sd
par(mfrow = c(1,1), mar = c(2,2, 1,1))
plot(predsd_ras, asp = 1, col = my.palette.var)
points(Fox_Point, pch = 16, cex = 0.5)
#plot(Scot_Shape_BNG, add = T)
```

