---
title: "trillium_phytoplasma"
author: "Julia Boyle"
date: "2023-05-27"
output: html_document
---
# Field data

## Load data and packages

```{r}
library(tidyverse)
library(cowplot)
library(patchwork)
library(car)
library(GLMMadaptive)
library(pscl)
library(lmerTest)
library(ggeffects)

trillium<-read.csv("Field_data/Trillium_Field_Data_Frequency.csv") %>% #just 2023
  mutate(Total_hemiptera_trillium=Young_hemiptera+Health_hemiptera+Infec_hemiptera)

trilliumY<-read.csv("Field_data/Trillium_Field_Data - Multiyear.csv") %>% #includes 2018 and 2023
 mutate(Year = as.factor(Year),
        Infection_presence=ifelse(Freq_infection_of_adults>0,"Yes","No"))

trillium_long<-trilliumY %>% 
  pivot_longer(cols=8:13,
  names_to = c("Status", "Hemiptera_presence"),
  names_sep = "_",
  values_to = "count"
) %>% 
  mutate(Infection=as.integer(ifelse(Status=="Infec", "1", "0")),
         Year=as.factor(Year))

str(trillium_long)

trillium_long_adults<-trillium_long %>% 
  filter(Status!="Young")

trillium_long

```

## Frequency of infection, density, and hemiptera

Zero-inflated poisson models looking at infection as a function of density and hemiptera frequency

```{r}

#2018 and 2023
##with random effects
head(trilliumY)
str(trilliumY)
MM4<-mixed_model(Total_infec_trilliums~Total_adult_trilliums+Year, random = ~ 1 | Population, data = trilliumY, family = zi.poisson(), zi_fixed = ~ Total_adult_trilliums+Year, zi_random=~1 |Population,control=list(iter_EM = 0))
summary(MM4) #loglik -254.7688. count model: density p=0.069. zero model:	 higher density increases likelihood of finding 0s p=0.01, higher year decreases chance of finding 0s=0.0388
##without random effects
summary(zeroinfl(Total_infec_trilliums~Total_adult_trilliums+Year, data=trilliumY)) #loglik -272.1, basically same results as above

#2023 only
##with random effects
head(trillium)
MM2<-mixed_model(Total_infec_trilliums~Freq_hemiptera_adults+Total_adult_trilliums, random = ~ 1 | Population, data = trillium, family = zi.poisson(), zi_fixed = ~ Freq_hemiptera_adults+Total_adult_trilliums, zi_random=~1 |Population,control=list(iter_EM = 0))
summary(MM2) #loglik -83.28, freq hemip 0.057 in count model 

##without random effects
summary(zeroinfl(Total_infec_trilliums~Freq_hemiptera_adults+Total_adult_trilliums, data=trillium)) #log lik -82.6, sig more likely to get zero infection in high adult density , marginal hemip in count model


#Association between trillium density and the proportion of trillium with hemiptera

# cor.test(trillium$Total_trilliums, trillium$Freq_hemiptera_total) #-0.475946
plot(trillium$Total_trilliums, trillium$Freq_hemiptera_total) 
mm1 = glm(Freq_hemiptera_total ~ Total_trilliums, weights = Total_trilliums, data = trillium, family = "binomial")
summary(mm1)

# Association between the frequency of hemiptera and frequency of infection (adults)

# cor.test(trillium$Freq_infection_of_adults, trillium$Freq_hemiptera_adults) #0.3341377
plot(trillium$Freq_hemiptera_adults, trillium$Freq_infection_of_adults) #0.3341377
mm2 = glm(Freq_infection_of_adults ~ Freq_hemiptera_adults, weights = Total_trilliums, data = trillium, family = "binomial")
summary(mm2)



```

## Figure 3
```{r}


fig3A<-trilliumY %>% 
  group_by(Total_adult_trilliums, Infection_presence, Year) %>% 
  tally() %>% 
  ggplot(aes(Total_adult_trilliums, fill = Infection_presence, colour = Infection_presence)) + 
  geom_density(alpha = 0.1)+
  facet_grid(~Year)+
  theme_cowplot()+
  xlab(expression('Density of flowering trillium (no./m'^2* ')'))+
  ylab("Density of sampled plots")+
  scale_fill_manual(name="Infection presence",values=c("grey", "darkgreen"))+
  scale_colour_manual(name="Infection presence",values=c("grey", "darkgreen"))+
  labs(subtitle = "Year p = 0.038, Trillium density p = 0.010")+ 
  # stats from model MM4
  theme(legend.position = "inside",legend.position.inside = c(0.25, 0.8)); fig3A

fig3B<-trilliumY %>% 
  group_by(Total_adult_trilliums, Total_infec_trilliums, Year) %>% 
  tally() %>% 
  ggplot(aes((Total_adult_trilliums), y=Total_infec_trilliums)) +
  geom_point(alpha=0.5)+
  facet_grid(~Year)+
  theme_cowplot()+
  xlab(expression('Density of flowering trillium (no./1 m'^2* ')'))+
  ylab("Number of infected trillium")+
  labs(subtitle = "Year p = 0.706, Trillium density p = 0.069"); fig3B
  # stats from model MM4


p1 = ggpredict(mm1, terms = "Total_trilliums[all]")
fig3C<- 
  ggplot()+
  geom_point(data = trillium, aes(x=Total_trilliums, y=Freq_hemiptera_total))+  
  geom_line(data = p1, aes(x = x, y = predicted), color="darkgreen") +
  geom_ribbon(data = p1, aes(x = x, ymin = conf.low, ymax = conf.high), fill="darkgreen", alpha = 0.2) +
  theme_cowplot()+
  xlab(expression('Density of all trillium (no./m'^2* ')'))+ 
  ylab("Proportion of all trillium with hemiptera")+
  labs(subtitle = "p < 0.001"); fig3C
# stats from model mm1


p2 = ggpredict(mm2, terms = "Freq_hemiptera_adults[all]")
fig3D<-
  ggplot() +
  geom_point(data = trillium, aes(x=Freq_hemiptera_adults, y=Freq_infection_of_adults)) +
  geom_line(data = p2, aes(x = x, y = predicted), color="darkgreen") +
  geom_ribbon(data = p2, aes(x = x, ymin = conf.low, ymax = conf.high), fill="darkgreen", alpha = 0.2) +
  xlab("Proportion of flowering trillium with hemiptera")+
  ylab("Proportion of flowering trillium infected")+ 
  theme_cowplot()+
  labs(subtitle = "p < 0.001")+
  ylim(0, NA); fig3D
# stats from model mm2

plot_grid(fig3A,fig3B,fig3C, fig3D, labels=c("A", "B", "C", "D"))  #10 by 10
ggsave("figures/Fig3.png", height = 10, width = 10)
ggsave("figures/Fig3.pdf", height = 10, width = 10)

```
Alternate Figure 3D (Figure S3)
```{r}
#testing that logit-transformed proportion data gives qualitatively the same result
cor.test(car::logit(trillium$Freq_hemiptera_adults, adjust = 0.001), car::logit(trillium$Freq_infection_of_adults, adjust = 0.001))
trillium %>%ggplot(aes(x=car::logit(Freq_hemiptera_adults, adjust = 0.001), y=car::logit(Freq_infection_of_adults, adjust = 0.001)))+geom_point()+geom_smooth(method="lm", color="darkgreen")+xlab("Logit-transformed proportion of flowering trillium with hemiptera")+ylab("Logit-transformed proportion of flowering trillium infected")+theme_cowplot()+labs(subtitle = "r = 0.260, p = 0.037")

trillium %>%
  ggplot(aes(x=Freq_hemiptera_adults, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="glm", formula = y ~ x, method.args = list(family = "binomial"), color="darkgreen")+xlab("Proportion of flowering trillium with hemiptera")+ylab("Proportion of flowering trillium infected")+theme_cowplot()
```


## Trait data
```{r}
traits<-read.csv("Trillium_Field_Data - Leaf and Height Phenotypes.csv")
str(traits)
traits<-traits %>% mutate(Infection_chr=ifelse(Infection=="0", "Asymptomatic", "Symptomatic"))
traits %>% group_by(Infection_chr) %>% tally()
```
```{r}
#Height
figheight<-traits %>% ggplot(aes(x=Infection_chr, y=Height_cm, colour=Infection_chr))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Height (cm)")+geom_point(alpha=0.2, position=position_jitterdodge(jitter.width=0.25, dodge.width=0.8))+scale_colour_manual(values=c("black", "darkgreen"))+theme(legend.position="none")
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Height_cm))
1-(23.34688/27.61000	)
27.61-23.34688

#leaf length
figleaf<-traits %>% ggplot(aes(x=Infection_chr, y=Leaf_length_cm, colour=Infection_chr))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Leaf length (cm)")+geom_point( position=position_jitterdodge(jitter.width=0.25, dodge.width=0.8), alpha=0.2)+scale_colour_manual(values=c("black", "darkgreen"))+theme(legend.position="none")

#sepal length
figsepal<-traits %>% ggplot(aes(x=Infection_chr, y=Sepal_length_cm, colour=Infection_chr))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Sepal length (cm)")+geom_point( position=position_jitterdodge(jitter.width=0.25, dodge.width=0.8), alpha=0.2)+scale_colour_manual(values=c("black", "darkgreen"))+theme(legend.position="none")
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Sepal_length_cm))
1-(3.460667/4.820938)

traits %>% ggplot(aes(x=Infection_chr, y=(Sepal_length_cm/Height_cm), colour=Infection_chr))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Ratio of sepal length to height")+geom_point( position=position_jitterdodge(jitter.width=0.25, dodge.width=0.8), alpha=0.2)+scale_colour_manual(values=c("black", "darkgreen"))+theme(legend.position="none")

#supplemental figure
plot_grid(figheight,figleaf,figsepal, labels=c("A", "B", "C")) #save 5 by 8
```
```{r}
hist(traits$Height_cm)
summary(lmer(Height_cm~Infection_chr+(1|Population), data=traits))
Anova(lmer(Height_cm~Infection_chr+(1|Population), data=traits), type=3)

hist(traits$Leaf_length_cm)
summary(lmer(Leaf_length_cm~Infection_chr+(1|Population), data=traits))
Anova(lmer(Leaf_length_cm~Infection_chr+(1|Population), data=traits), type=3) #not sig

hist(log(traits$Sepal_length_cm))
shapiro.test(log(traits$Sepal_length_cm))
summary(lmer(log(Sepal_length_cm)~Infection_chr+(1|Population), data=traits))
Anova(lmer(log(Sepal_length_cm)~Infection_chr+(1|Population), data=traits), type=3)
```

## Seed counts 2024
```{r}
#actual seed counts only
seed<-read.csv("2024 Trillium seed counts and predicted counts.csv")
head(seed)
seed$seed_count_round<-round(seed$seed_count)
seed<-seed %>% mutate(Full_treatment=paste(Infection_status, Seed_count_real_or_predicted, sep = '_'))

seed %>% drop_na(seed_count) %>%  ggplot(aes(x=Infection_status, y=seed_count, colour=Seed_count_real_or_predicted))+geom_boxplot()+geom_point( position=position_jitterdodge(jitter.width=0.25, dodge.width=0.8), alpha=0.2)+theme_cowplot()+ylab("Seed count")+xlab("Infection status")+scale_x_discrete(labels=c("Asymptomatic", "Symptomatic"))+scale_colour_manual(values=c("goldenrod", "darkgrey"),name="Seed count", labels=c("Predicted from leaf length", "Real"))+theme(legend.position="top")

seed_tril_plot<-seed %>% drop_na(seed_count) %>%filter(Full_treatment!="Nonsymptomatic_Real") %>%   ggplot(aes(x=Infection_status, y=seed_count, colour=Seed_count_real_or_predicted))+geom_boxplot()+geom_point( position=position_jitterdodge(jitter.width=0.25, dodge.width=0.8), alpha=0.2)+theme_cowplot()+ylab("Seed count")+xlab("Infection status")+scale_x_discrete(labels=c("Asymptomatic", "Symptomatic"))+scale_colour_manual(values=c("black", "darkgreen"),name="Seed count", labels=c("Modeled", "Counted"))+theme(legend.position="top")

seed %>%filter(Paired_data_plants=='Yes'& Seed_count_real_or_predicted=='Real') %>%  drop_na(seed_count)%>% group_by(Infection_status) %>% tally() #only 81 plants had real seed count numbers
seed %>%filter(Paired_data_plants=='Yes') %>% group_by(Infection_status) %>% filter(seed_count==0) %>% tally()
51/81
seed_comparison1<-seed %>% filter(Paired_data_plants=='Yes'&Infection_status=="Symptomatic")
seed_comparison2<-seed %>% filter(Full_treatment=="Symptomatic_Real" | Full_treatment== "Nonsymptomatic_Predicted_from_lf_length")
hist((seed_comparison2$seed_count))
summary(glmer.nb(round(seed_count)~Seed_count_real_or_predicted+(1|pair_number), data=seed_comparison1))
Anova(glmer.nb(round(seed_count)~Seed_count_real_or_predicted+(1|pair_number), data=seed_comparison1), type=3) #comparing infected observed vs. expected seed number based on size
summary(glmer.nb(round(seed_count)~Infection_status+(1|pair_number), data=seed_comparison2, control = glmerControl("bobyqa")))
(Anova(glmer.nb(round(seed_count)~Infection_status+(1|pair_number), data=seed_comparison2, control = glmerControl("bobyqa")), type=3)) #comparing the focal paired plants, actual infected and predicted seed count of healthy paired plant

#try quasipoisson if overdispersed
plyr::ddply(seed_comparison2, c("Infection_status"), summarise,
                  N=sum(!is.na(seed_count)),
             mean = mean ((seed_count), na.rm=TRUE),
             sd = sd (seed_count,na.rm=TRUE),
             se = sd/sqrt(N))
((5.851852		- 18.525928	)/18.525928)*100
plyr::ddply(seed_comparison1, c("Seed_count_real_or_predicted"), summarise,
                  N=sum(!is.na(seed_count)),
             mean = mean ((seed_count), na.rm=TRUE),
             sd = sd (seed_count,na.rm=TRUE),
             se = sd/sqrt(N))
((5.851852		- 16.522099	)/16.522099)*100
```

```{r}
all_data <-read.csv('Trillium_data_collection - 2024 demography.csv')

all_data <- all_data %>%
  filter(!is.na(Date))

#adding a column to make D=1, H=0
all_data$DH_number <- all_data$DH_type
all_data$DH_number[all_data$DH_number == "D"] <- 1
all_data$DH_number[all_data$DH_number == "H"] <- 0
all_data$DH_number = as.numeric(all_data$DH_number)

# Demography data from 2024 with more age classes
#juvenile count= no. plants with 3 leaves but no flower
#adult count = no. plants with 3 leaves and flower but no infection
#infected adult count = no. plants with 3 leaves and flower and infection
#one leaf count = no. plants at 1 leaf stage

all_data<-all_data %>% mutate(Total_trilliums=(infected_adult+adult_count+juvenile_count+one_leaf_count+cot_count)) %>% mutate(Freq_infected_adult=infected_adult/(infected_adult+adult_count)) %>% mutate(count_3leaved=(juvenile_count+infected_adult+adult_count))
all_data<-all_data %>% mutate(Ratio_one_to_repro=(one_leaf_count/(infected_adult+adult_count)))
all_data<-all_data %>% mutate(Ratio_one_to_three=(one_leaf_count/(infected_adult+adult_count+juvenile_count)))
all_data<-all_data %>% mutate(Ratio_veg_to_repro=(juvenile_count/(infected_adult+adult_count)))
#all_data$Freq_infected_adult[is.na(all_data$Freq_infected_adult)] <- 0 
#all_data$Freq_infected_adult[is.nan(all_data$Freq_infected_adult)] <- 0 
all_data<-all_data %>% mutate(across(where(is.numeric), \(x) ifelse(is.nan(x), NA, x)))
all_data$Ratio_one_to_repro[is.infinite(all_data$Ratio_one_to_repro)] <- NA
all_data$Ratio_veg_to_repro[is.infinite(all_data$Ratio_veg_to_repro)] <- NA

cor.test(all_data$Freq_infected_adult, all_data$Ratio_one_to_repro)
library(bestNormalize)
hist((all_data$Ratio_one_to_repro))
hist(predict(bestNormalize(all_data$Ratio_one_to_repro)))
demography2024_one<-ggplot(data=all_data, aes(Freq_infected_adult, Ratio_one_to_repro))+geom_point()+geom_smooth(method="lm", color="darkgreen")+theme_cowplot()+ylab("One-leaved:flowering ratio")+xlab("Proportion of flowering trillium infected")+labs(subtitle = "r = -0.082, p = 0.336")

cor.test(all_data$Freq_infected_adult, all_data$Ratio_veg_to_repro)
demography_2024_veg<-all_data %>% ggplot(aes(Freq_infected_adult, Ratio_veg_to_repro))+geom_point()+geom_smooth(method="lm", color="darkgreen")+theme_cowplot()+ylab("Three-leaved non-flowering:flowering ratio")+xlab("Proportion of flowering trillium infected")+ggtitle("2024")+labs(subtitle = "r = -0.079, p = 0.352")

cor.test(all_data$Freq_infected_adult, all_data$Ratio_one_to_three)
ggplot(data=all_data, aes(Freq_infected_adult, Ratio_one_to_three))+geom_point()+geom_smooth(method="lm", color="darkgreen")+theme_cowplot()+ylab("One-leaved:Three-leaved ratio")+xlab("Proportion of flowering trillium infected")+labs(subtitle = "r = -0.023, p = 0.784")
```

## Figure 2
```{r}



#Association between adult infection frequencies and the ratio of young plants to adults

cor.test(trillium$Freq_infection_of_adults, trillium$Young_to_repro_ratio) #0.1034085
plot(trillium$Freq_infection_of_adults, trillium$Young_to_repro_ratio) #0.1034085
mm3 = lm(log(Young_to_repro_ratio) ~ Freq_infection_of_adults, data = trillium)
summary(mm3)

#making the demography data plot from 2023
cor.test(trillium$Freq_infection_of_adults, trillium$Young_to_repro_ratio)
demography2023<-trillium %>% 
  ggplot(aes(x=Freq_infection_of_adults, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm", color="darkgreen")+theme_cowplot()+xlab("Proportion of flowering trillium infected")+ylab("Three-leaved non-flowering:flowering ratio")+labs(subtitle = "r = 0.103, p = 0.412")+ggtitle("2023")
demography2023
```
```{r}
#plot_grid(seed_tril_plot, demography2023, labels=c("A", "B")) #5 by 8 landscape for ellie's version
#5 by 10 for the above version

plot_grid(seed_tril_plot, demography2024_one, labels=c("A", "B")) #5 by 10.5

#plot_grid(seed_tril_plot, demography2023, demography2024_one, demography_2024_veg, labels="AUTO") #8 by 10
```

# iNaturalist data
```{r}
#install.packages('geodata')
library(maps)
library(mapdata)
library(ggmap)
library(RColorBrewer)
library(raster)
library(scales)
library(CoordinateCleaner)
library(rnaturalearth)
library(rnaturalearthdata)
library(geodata)
library(sf)
library(plyr)
library(spData)
library(tidyterra)
library(terra)
library(tmap) 
library(predicts)
library(tidyverse)
library(cowplot)
```

```{r}
#inat<-read.csv("iNat_alltrilliums_2023_07_05.csv") #only up until 2023
inat<-read.csv("iNat_alltrilliums_2025_03_14_539340.csv") #version with 2024 data
head(inat)

set.seed(2025)
inat_cleaned = inat %>% 
  # Adding positional accuracy filter to see if that helps deal with observations in lakes
  filter(!is.na(latitude), !is.na(longitude), !is.na(public_positional_accuracy)) %>% 
  filter(public_positional_accuracy < 10000) %>% 
  dplyr::mutate(infected = `field.phytoplasma.infection.`) %>% 
  mutate(infected = if_else(is.na(infected), "No", infected)) %>% 
  # Drop european and puerto rico observations
  filter(latitude > 20, longitude < -65) %>%  #at this point 32,649 observations
  # Group by user and subset to 15 observations per user
  group_by(user_login) %>% 
  slice_sample(n = 15, replace = FALSE) #29,412 observations

inat_clean <- clean_coordinates(x = inat_cleaned, 
                           lon = "longitude", 
                           lat = "latitude",
                           species = "scientific_name",
                           tests = c("centroids",
                                    "equal", "zeros","institutions", "duplicates")) #28,371 observations
inat_clean=inat_clean %>% filter(.summary==TRUE) #only include observations that have no suspect issues
inat_clean=inat_clean %>% mutate(infected=ifelse(grepl("Yes", infected), "1", "0"))
inat_clean$infected

inat%>% filter( field.phytoplasma.infection.=="Yes") #846
inat_clean%>% filter( field.phytoplasma.infection.=="Yes") #526 after filtering out stuff with clean_coordinates
```
```{r}
inat_clean_tally<-inat_clean %>% group_by(user_login) %>% tally()
mean(inat_clean_tally$n)
median(inat_clean_tally$n)

inatinfec<-inat_clean %>% filter( field.phytoplasma.infection.=="Yes") #make another dataset of infected points
inatinfec %>% group_by(user_login) %>% tally() #vast majority of users contributed less than 6 infected observations, one person contributed 12
inat_clean %>% group_by(user_login) %>% tally()

inat_clean %>% filter(observed_on>'2020-01-01') #23,385
23385/28371

inat_clean %>% filter(observed_on>'2025-01-01')
```

## Extracting coordinate data
Run this everytime you load R because spatrasters do not save with R enviro
```{r}
dir.create(path = "data") 
dir.create(path = "output")
bioclim_data_source <- worldclim_global(var = "bio", res = 0.5, path = "data/")
#Determine geographi cextent of our data 
max_lat<-ceiling(max(inat_clean$latitude)) 
min_lat<-floor(min(inat_clean$latitude)) 
max_lon<-ceiling(max(inat_clean$longitude)) 
min_lon<-floor(min(inat_clean$longitude)) 
#Store boundaries in a single extent object 
geographic_extent<-ext(x=c(min_lon,max_lon,min_lat,max_lat))

# Make an extent that is 25% larger 
sample_extent <- geographic_extent * 1.25 
# Crop bioclim data to desired extent 
bioclim_data <- crop(x = bioclim_data_source, y = sample_extent) 
# Plot the first of the bioclim variables to check on cropping 
plot(bioclim_data[[1]])
plot(bioclim_data[[12]])

bioclim_extract <- terra::extract(x = bioclim_data, y = as.matrix(inat_clean[, c("longitude", "latitude")])) # No need for an ID column
# I GET 41 OBSERVATIONS WITH NAS FROM BIOCLIM DATA- WHY???
# Add the point and climate datasets together 
points_climate <- cbind(inat_clean, bioclim_extract)

crop_data<-cropland(source="WorldCover",  path = "data/")
crop_data<- crop(x = crop_data, y = sample_extent) 
plot(crop_data[[1]])
cropdata_extract<- terra::extract(x = crop_data, y = as.matrix(inat_clean[, c("longitude", "latitude")]))
points <- cbind(points_climate, cropdata_extract)

#stonefruit<-crop_monfreda(crop="stonefruitnes", var="area_ha",  path = "data/") #download failed?

footprint<-footprint(year=2009, path = "data/")
footprint_crop<- crop(x = footprint, y = sample_extent) 
plot(footprint_crop[[1]])
names(footprint_crop)<-"footprint"
footprint_extract<- terra::extract(x = footprint_crop, y = as.matrix(inat_clean[, c("longitude", "latitude")]))
#footprint_extract<-rename(footprint_extract, footprint = `wildareas-v3-2009-human-footprint`)
points<-cbind(points, footprint_extract)

bioclim_data
footprint_crop
crop_data
geo_data<-c(bioclim_data, footprint_crop, crop_data)
geo_data_test<-geo_data
```

##Density map
```{r}
dat_sf = st_as_sf(inat_clean, coords = c("longitude", "latitude")) %>% 
  st_set_crs("EPSG:4326")
# Reproject to equal area projection
dat_ea = st_transform(dat_sf, crs = "ESRI:102003")
# Make a grid of polygons that encompass the observation points
# Choosing an arbitrary 5 km grid size
area_fishnet_grid = st_make_grid(dat_ea, c(5e4, 5e4), what = "polygons", square = FALSE)

# Convert to sf and add grid ID
fishnet_grid_sf = st_sf(area_fishnet_grid) %>%
  # add grid ID
  mutate(grid_id = 1:length(lengths(area_fishnet_grid)))
# Assign observations to grid cells
grid_assignments = st_intersection(fishnet_grid_sf, dat_ea)

# Calculate frequencies per grid cell
freq_per_poly = grid_assignments %>% 
  group_by(grid_id) %>% 
  summarize(n = n(), n_infected = sum(infected == "1")) %>% 
  mutate(freq_infected = n_infected/n)

# Rejoin these calculated frequencies to polygons for plotting
freqs_with_polys = st_join(fishnet_grid_sf, freq_per_poly) %>% 
  filter(!is.na(n))
plot(freqs_with_polys)
freqs_with_polys$n
freqs_with_polys$freq_infected

world = ne_countries(scale = "large", returnclass = "sf", continent = "north america")
lakes = ne_download(scale = "medium", type = "lakes", category = "physical")

tril_b<-ggplot() +
  geom_sf(data = world, fill = "white") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  geom_sf(data = freqs_with_polys, aes(fill = freq_infected), alpha = 0.8, color = "lightgrey") +
  scale_fill_gradientn(colours = c("white", "#33FF99","#00CC66","black"), name="Proportion \ninfected") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()+ theme(legend.position = c(0.15, 0.23))+xlab("Longitude")+ylab("Latitude")

tril_a<-ggplot() +
  geom_sf(data = world, fill = "white") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  geom_sf(data = (freqs_with_polys %>% filter(n>0)), aes(fill = as.numeric(n)), alpha = 0.8, color = "lightgrey") +
  scale_fill_gradientn(colours = c( "#CCFFFF","#3399FF", "navy"), name="Observations") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()+ theme(legend.position = c(0.15, 0.23))+xlab("Longitude")+ylab("Latitude")

cowplot::plot_grid(tril_a, tril_b, labels=c("A", "B"), ncol=1) #10 by 7 portrait
trilab<-plot_grid(tril_a, tril_b, labels=c("A", "B"), nrow=1) #5 by 11 landscape
```
##Moran’s I analysis
https://mgimond.github.io/simple_moransI_example/
```{r}
library(sf)
library(spdep)
library(tmap)
nb <- poly2nb(freqs_with_polys, queen=TRUE)
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
inc.lag <- lag.listw(lw, freqs_with_polys$n_infected)
plot(inc.lag ~ freqs_with_polys$n_infected, pch=16, asp=1)
M1 <- lm(inc.lag ~ freqs_with_polys$n_infected)
abline(M1, col="blue")
coef(M1)[2] #0.4040173
#this slope is Morans I: because this is positive, it means polygons with high frequency of infection are more likely to have neighbouring polygons with high frequency of infection
#a positive Moran’s I value suggests clustering whereas a negative Moran’s I value suggests dispersion
inc.lag.freq <- lag.listw(lw, freqs_with_polys$freq_infected)
plot(inc.lag.freq ~ freqs_with_polys$freq_infected, pch=16, asp=1)
M1.freq <- lm(inc.lag.freq ~ freqs_with_polys$freq_infected)
abline(M1.freq, col="blue")
coef(M1.freq)[2] #0.1074369

moran(freqs_with_polys$n_infected, lw, length(nb), Szero(lw))[1] #0.4122603, infections are pretty clustered in space
moran(freqs_with_polys$freq_infected, lw, length(nb), Szero(lw))[1] #0.1084906
moran(freqs_with_polys$n, lw, length(nb), Szero(lw))[1] #0.3111099

MC.ninfec<- moran.mc(freqs_with_polys$n_infected, lw, nsim=999, alternative="greater")
MC.ninfec
plot(MC.ninfec)
MC.freq<- moran.mc(freqs_with_polys$freq_infected, lw, nsim=999, alternative="greater")
MC.freq
plot(MC.freq)
MC.n<- moran.mc(freqs_with_polys$n, lw, nsim=999, alternative="greater")
MC.n
plot(MC.n)
```

##Models
what seems to cause infection?
```{r}
#as.data.frame(grid_assignments)
#as.data.frame(freq_per_poly)
inat_grid<-merge(as.data.frame(grid_assignments), as.data.frame(freq_per_poly), by.x="grid_id", by.y="grid_id")
inat_grid<-inat_grid %>% dplyr::select(grid_id, id, n, n_infected, freq_infected)
inat_full<-merge(inat_grid, points, by.x="id", by.y="id")
inat_full$infected<-as.numeric(inat_full$infected)
#inat_full$footprint<-inat_full$`wildareas-v3-2009-human-footprint`
inat_full$Year<-substring(inat_full$observed_on,1,4)
#str(inat_full)
```
```{r}
#add total trillium number as a layer to the raster 
r <- rast(xmin=-100.875, xmax=-62.125, ymin=31.125, ymax=49.875, ncols=4650, nrows=2250, crs = crs(geo_data_test))
# you need to provide a function 'fun' for when there are multiple points per cell
inat_sub_matrix<-as.matrix(inat_full%>% select(longitude, latitude,n))
trillium_number <- rasterize(inat_sub_matrix[,1:2], r, inat_sub_matrix[,3], fun=max)
trillium_number_resampled <- resample(trillium_number, geo_data_test)
names(trillium_number_resampled)<-"n"
plot(trillium_number_resampled)
geo_data_test<-c(geo_data_test, trillium_number_resampled)

points<-SpatialPoints(inat_sub_matrix[, c('longitude','latitude')], inat_sub_matrix[, c('n')])
pixels<-SpatialPixelsDataFrame(points, tolerance=0.916421, points@data)
raster<-raster(pixels[,'n'])

#add lat and long as a layer to the raster??
r <- rast(xmin=-100.875, xmax=-62.125, ymin=31.125, ymax=49.875, ncols=4650, nrows=2250, crs = crs(geo_data_test))
longitude<-latitude<-r
longitude[]<-xFromCell(r, 1:ncell(r))
names(longitude)<-"longitude"
latitude[]<-yFromCell(r, 1:ncell(r))
names(latitude)<-"latitude"
plot(longitude)  
plot(latitude)  
geo_data_test<-c(geo_data_test, longitude)
geo_data_test<-c(geo_data_test, latitude)
names(geo_data_test)
#cellFromXY(r)
```

### INLA
https://ourcodingclub.github.io/tutorials/inla/
https://ourcodingclub.github.io/tutorials/spatial-modelling-inla/
```{r}
set.seed(2025)
#install.packages("devtools")
#if(!require(ggregplot)) devtools::install_github("gfalbery/ggregplot") # Installing Greg's package for plotting functions!
#install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA); library(ggregplot); library(RColorBrewer)
library(INLA)
library(INLAspacetime)
library(usdm)
head(inat_full)

##########################################################################
which(is.na(inat_full$wc2.1_30s_bio_1)) #41 uninfected trillium near quebec and montreal don't have associated bioclim data for some reason. we need to exclude them for INLA
inat_full[which(is.na(inat_full$wc2.1_30s_bio_1)),]
inat_full_complete = inat_full[complete.cases(inat_full[53:73]),]
##########################################################################

phen <- c("longitude", "latitude") # Base columns with spatial information we'll need

resp <- "infected" # Response variable

covar <- c("n","wc2.1_30s_bio_1",  "wc2.1_30s_bio_2","wc2.1_30s_bio_3","wc2.1_30s_bio_4","wc2.1_30s_bio_5","wc2.1_30s_bio_6","wc2.1_30s_bio_7","wc2.1_30s_bio_8","wc2.1_30s_bio_9","wc2.1_30s_bio_10","wc2.1_30s_bio_11","wc2.1_30s_bio_12","wc2.1_30s_bio_13","wc2.1_30s_bio_14","wc2.1_30s_bio_15","wc2.1_30s_bio_16","wc2.1_30s_bio_17","wc2.1_30s_bio_18","wc2.1_30s_bio_19","cropland", "footprint")

#Selecting final variables for the model
ggregplot::INLAModelSel(resp, covar, Family="binomial", Data=inat_full_complete) #removes variables that improve DIC by at least 2
  #[1] "Losing wc2.1_30s_bio_18"[1] "Run 2"[1] "Losing wc2.1_30s_bio_1"[1] "Run 3"[1] "Losing wc2.1_30s_bio_13"[1] "Run 4" "Losing wc2.1_30s_bio_16"[1] "Run 5[1]   "Losing wc2.1_30s_bio_17"[1] "Run 6"[1] "Losing n"[1] "Run 7"[1] "Losing wc2.1_30s_bio_5"[1] "Run 8"[1] "Losing footprint"[1] "Run 9"[1] "Losing   wc2.1_30s_bio_11"[1] "Run 10"[1] "Losing wc2.1_30s_bio_15"[1] "Run 11" 
  #it removed n and footprint, but I want to keep them in the model due to a priori reasons

vifcor(inat_full_complete %>% dplyr::select(n,wc2.1_30s_bio_2,wc2.1_30s_bio_3,wc2.1_30s_bio_4,wc2.1_30s_bio_6,wc2.1_30s_bio_7,wc2.1_30s_bio_8,wc2.1_30s_bio_9,wc2.1_30s_bio_10,wc2.1_30s_bio_12,wc2.1_30s_bio_14,wc2.1_30s_bio_19,cropland, footprint)) #remove vars 19, 7, 4, 12 due to multicollinearity #except var 3 still has REALLY HIGH VIF! Let's see what happens when we remove it
vifcor(inat_full_complete %>% dplyr::select(n,wc2.1_30s_bio_2,wc2.1_30s_bio_6,wc2.1_30s_bio_8,wc2.1_30s_bio_9,wc2.1_30s_bio_10,wc2.1_30s_bio_14,cropland, footprint))#After removing var 3, all VIFs are under 5, which is great

Finalcovar<- c("n","wc2.1_30s_bio_2","wc2.1_30s_bio_6","wc2.1_30s_bio_8","wc2.1_30s_bio_9","wc2.1_30s_bio_10","wc2.1_30s_bio_14","cropland", "footprint")

#incorporating space
Locations = cbind(inat_full_complete$longitude, inat_full_complete$latitude) # using the sampling locations
Mesh <- INLA::inla.mesh.2d(Locations, max.edge = c(10, 500), cutoff=1) #smaller triangles- use smaller triangles for publication
plot(Mesh)
points(Locations, col = "red", pch = 2)
save(Mesh, file = "trillium_inla_mesh.Rdata") #!!!
# Making the A matrix
HostsA <- INLA::inla.spde.make.A(Mesh, loc = Locations) # Making A matrix, which essentially translates spatial locations on the mesh into vectors in the model.
Range_prior<-inat_full_complete[, 28:29]%>% slice(1:n(), 100)%>% dist%>% c%>% max%>% divide_by(2) #14.72662.
Range_prior<-14.72662
#This is the null range prior. This method of determining the prior is found from code in https://doi.org/10.1098/rspb.2020.2655
Hosts.spde = INLA::inla.spde2.pcmatern(mesh = Mesh, prior.range = c(Range_prior, 0.5), prior.sigma = c(.5, .5)) 
w.Host <- INLA::inla.spde.make.index('w', n.spde = Hosts.spde$n.spde) # making the w

# Making the model matrix #### 
X0 <- model.matrix(as.formula(paste0("~ -1 +", paste(Finalcovar, collapse = "+"))), data = inat_full_complete) # make the model matrix using the final model selection formula without a response variable.
X <- as.data.frame(X0)
# Making the stack ####
N <- nrow(inat_full_complete)
StackHost <- INLA::inla.stack(
  data = list(y = inat_full_complete[,resp]), # specify the response variable
  A = list(1,1,HostsA), # Vector of Multiplication factors for random and fixed effects  #put as many 1s as there are fixed and random effects in the effects() below            
  effects = list(
    Intercept = rep(1, N), # specify the manual intercept!
    X = X, # attach the model matrix
    # insert vectors of any random effects (none here)
    w = w.Host)) # attach the w 
#Running the model
f1 <- as.formula(paste0("y ~ -1 + Intercept + ", paste0(colnames(X), collapse = " + "))) #version only fixed effects
f3 <- as.formula(paste0("y ~ -1 + Intercept + ", paste0(colnames(X), collapse = " + "), " + f(w, model = Hosts.spde)")) #version with space SPDE random effect
IM1 <- INLA::inla(f1, # Base model (no random effects)
            family = "binomial",
            data = inla.stack.data(StackHost),
            control.compute = list(dic = TRUE),
            control.predictor = list(A = INLA::inla.stack.A(StackHost)))
IM3 <- INLA::inla(f3, # SPDE random effect 
            family = "binomial",
            data = inla.stack.data(StackHost),
            control.compute = list(dic = TRUE),
            control.predictor = list(A = INLA::inla.stack.A(StackHost)))
IM3.r<-inla.rerun(IM3) #rerun to make sure values are as stable as possible
SpatialHostList <- list(IM1, IM3.r) 

sapply(SpatialHostList, function(f) f$dic$dic) #does space make it better?
4705.980-4380.141 #Spatial model is better by DIC 325.84!
ggregplot::INLADICFig(SpatialHostList, ModelNames = c("Base", "SPDE")) #if DIC is smaller by >2, then suggests autocorrelation was important to include
dput(summary(IM3.r), file="INLAsummary_IM3.txt", control="all")
summary(IM3.r)#There are no P values in INLA. Importance or significance of variables can be deduced by examining the overlap of their 2.5% and 97.5% posterior estimates with zero.
summary(SpatialHostList[[1]]) #base model summary

IM3.r$summary.fitted.values #the predicted infection
IM3.r$summary.random #this is how the random effects (n=205) affected likelihood of an observation being infected
IM3.r$summary.hyperpar #overall summary of random effects of space
IM3.r$summary.fixed #this is how the fixed effects (n=10) affected likelihood of an observation being infected

#saveRDS(SpatialHostList, paste0("output/","SpatialHostList",".rds"))
#saveRDS(IM3.r, paste0("output/","IM3.r",".rds"))
```

####Visuals
Model significances
```{r}
tril_d<-ggregplot::Efxplot(SpatialHostList)+scale_colour_discrete(labels=c("Spatial", "Base"))+cowplot::theme_cowplot()+
  cowplot::background_grid(major = "y", minor = "y", size.minor = 0.5)+scale_x_discrete(
    "Fixed effect",
    labels = c(
      "Intercept" = "Intercept",
      "n" = "Observation density",
      "wc2.1_30s_bio_2" = "Mean diurnal range",
      "wc2.1_30s_bio_6" = "Min. temp. of coldest month",
      "wc2.1_30s_bio_8" = "Mean temp. of wettest quart.",
      "wc2.1_30s_bio_9" = "Mean temp. of driest quart.",
      "wc2.1_30s_bio_10" = "Mean temp of warmest quart.",
      "wc2.1_30s_bio_14" = "Precip. of driest month", 
      "cropland" = "Cropland",
      "footprint" = "Footprint" ))+scale_colour_manual(labels=c("SPDE","Base"),values=c("darkgreen", "grey"))
tril_d
```

Spatial fields
```{r}
inla.spde2.result(inla=IM3.r, name="w", spde=Hosts.spde, do.transform = T) #extracting info on the spatial field, while transforming results back from internal model scale
#the higher the Kappa, the smoother the spatial autocorrelation structure. Shorter range and smaller kappa indicate higher autocorrelation between closely related points
```

```{r}
set.seed(2025)
usa <- map_data("state")
canada <- map_data("worldHires","Canada")

nslbls <- unlist(lapply(nsbrks, function(x) ifelse(x < 0, paste(x, "°S"), ifelse(x > 0, paste(x, "°N"),x))))

tril_c<-ggregplot::ggField(IM3.r, Mesh, Groups = 1) +
  scale_fill_brewer(palette = "Greens", name="Random \nEffect \nMean")+ geom_polygon(data = usa, 
                                 aes(x=long, y = lat, group = group), 
                                 fill = NA, 
                                 color="black") +
    geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
                 fill = NA, color="black")+
    coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE)+xlab("Longitude")+ylab("Latitude")+theme_bw()+scale_x_discrete(labels = c(
      "-100" = "100°W",
      "-95" = "95°W",
      "-90" = "90°W",
      "-85" = "85°W",
      "-80" = "80°W",
      "-75" = "75°W",
      "-70" = "70°W"))#scale_y_discrete(labels = c("30" = "30°N","35" = "35°N","40" = "40°N","45" = "45°N","50" = "50°N","55" = "55°N"))
```

####Figure 4
```{r}
plot_grid(tril_a, tril_b, tril_c, tril_d, labels="AUTO") #12 by 10 landscape
```

Spatial field standard deviation
```{r}
#I modified the code from the package to get a version that gives me sd instead of mean. The function is on github as ggField_sd.R
spatial_field_sd<-ggField_sd(IM3.r, Mesh, Groups = 1) +
  scale_fill_brewer(palette = "Reds")+ geom_polygon(data = usa, 
                                 aes(x=long, y = lat, group = group), 
                                 fill = NA, 
                                 color="black") +
    geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
                 fill = NA, color="black")+
    coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE)+xlab("Longitude")+ylab("Latitude")+theme_bw()+scale_x_discrete(labels = c(
      "-100" = "100°W",
      "-95" = "95°W",
      "-90" = "90°W",
      "-85" = "85°W",
      "-80" = "80°W",
      "-75" = "75°W",
      "-70" = "70°W"))

plot_grid(tril_c, spatial_field_sd, labels="AUTO") #save 3 by 10 landscape
```

```{r}
library(inlabru) 
#masking the regions of water requires I have the boundaries as a spatialpolygons object
ggplot()+gg(Mesh, color=IM3.r$summary.random$w$mean, mask=world_trans, nx=300, ny=300) #, mask=(usa+canada) #world or lakes
world_trans<-st_transform(world, "ESRI:102003")
ggplot()+gg(Mesh, color=IM3.r$summary.random$w$mean)+
  geom_sf(data = world, fill = "NA") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()#+ theme(legend.position = c(0.15, 0.23))
```
