---
title: "Trillium iNaturalist data"
author: "Julia Boyle"
date: "2024-02-12"
output: html_document
---

TO DO:
-Turn lat long into equal area projection (eg. 1km by 1km squares), so we have gridded cells. Look at proportion infected to total observations in each cell.
-Look at year by year (or binned time periods) proportion infected individuals 
-Add user ID as random effect or limit number of observations per user? Or see what the max number of infected observations there are by user ID to see if anyone was specifically seeking them out or not, useful to report
-occupancy model for space: different number of samples per grid, accounting for them
-calculate infection frequency (prevalence and presence) and trillium density, then see whether they are correlated
-correlate infection with climate data?
-correlate infection with agricultural/ insect vector/ other plant host data?


# iNaturalist data
```{r}
remotes::install_github("dkahle/ggmap")
library(tidyverse)
library(cowplot)
library(patchwork)
library(maps)
library(mapdata)
library(ggmap)
library(RColorBrewer)

#library(maptools)
#library(rnaturalearth)
#library(rnaturalearthdata)
#library(rnaturalearthhires)
#library(raster)

inat<-read.csv("iNat_alltrilliums_2023_07_05.csv") #full dataset, downloaded july 5th 2023
head(inat)

#time series
inat20s<-inat %>% filter(observed_on>2020-1-1, observed_on<2030-1-1)
inat10s<-inat %>% filter(observed_on>2010-1-1, observed_on<2020-1-1)
inat00s<-inat %>% filter(observed_on>2000-1-1, observed_on<2010-1-1)
inat90s<-inat %>% filter(observed_on>1990-1-1, observed_on<2000-1-1)
inat80sandbelow<-inat %>% filter(observed_on<1990-1-1)

#infected time series
inat20s_infec<-inat20s %>% filter( field.phytoplasma.infection.=="Yes")
inat10s_infec<-inat10s %>% filter( field.phytoplasma.infection.=="Yes")
inat00s_infec<-inat00s %>% filter( field.phytoplasma.infection.=="Yes")
inat90s_infec<-inat90s %>% filter( field.phytoplasma.infection.=="Yes")
inat80sandbelow_infec<-inat80sandbelow %>% filter( field.phytoplasma.infection.=="Yes")

#infected observations only
inatinfec<-inat %>% filter( field.phytoplasma.infection.=="Yes")
```

# Bias correction
Are any users contributing way more than others to the dataset?
```{r}
inat %>% ggplot(aes(x=user_login, fill=id))+geom_bar()+theme(axis.text.x = element_text(size=10, angle = 45, hjust = 1))
inat %>%  filter( field.phytoplasma.infection.=="Yes") %>% 
  ggplot(aes(x=user_login, fill=id))+geom_bar()+theme(axis.text.x = element_text(size=10, angle = 45, hjust = 1))

inat %>%  filter( field.phytoplasma.infection.=="Yes") %>% group_by(user_login) %>% tally()

inat %>% group_by(user_login) %>% tally()

inat %>% filter(user_login=="jemredwood")
```

# Visual maps

could try having size of bubbles indicating number of infected https://stackoverflow.com/questions/60021868/producing-heat-map-over-geo-locations-in-r 

## Observations are points
```{r}
#http://mbontrager.org/blog/2016/08/01/Maps-in-R
infectedpoints <- filter(inat, field.phytoplasma.infection.=="Yes")
normalpoints <- filter(inat, field.phytoplasma.infection.=="")
infcolour <- brewer.pal(6, "Paired")[5]
normcolour <- brewer.pal(6, "Paired")[3]

usa <- map_data("usa")
canada <- map_data("worldHires","Canada")
NAmap <- ggplot() + geom_polygon(data = usa, 
                                 aes(x=long, y = lat, group = group), 
                                 fill = "white", 
                                 color="black") +
    geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
                 fill = "white", color="black") +
    coord_fixed(xlim = c(-95, -70),  ylim = c(33, 50), ratio = 1.2)
NAmap

range<-NAmap+ geom_point(data=inat, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 1967-2023")
infections<-NAmap+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 1989-2023")
range+infections
```
Time series
```{r}
#2020 and above
NAmap+ geom_point(data=inat20s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 2020-2023")
NAmap+ geom_point(data=inat20s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 2020-2023")

#2010-2020
NAmap+ geom_point(data=inat10s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 2010-2020")
NAmap+ geom_point(data=inat10s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 2010-2020")

#2000-2010
NAmap+ geom_point(data=inat00s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 2000-2010")
NAmap+ geom_point(data=inat00s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 2000-2010")

#1990-2000
NAmap+ geom_point(data=inat90s, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 1990-2000")
NAmap+ geom_point(data=inat90s_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 1990-2000")

#1980s and below
NAmap+ geom_point(data=inat80sandbelow, aes(x=longitude, y=latitude), fill="blue",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Trillium grandiflorum range 1980s and below")
NAmap+ geom_point(data=inat80sandbelow_infec, aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")+ggtitle("Phytoplasma infected Trillium grandiflorum 1980s and below")
```

## Heat map
```{r}
#https://axelhodler.medium.com/creating-a-heat-map-from-coordinates-using-r-780db4901075
map_bounds <- c(left = -90, bottom = 30, right = -70, top = 50)
coords.map <- get_stamenmap(map_bounds, zoom = 7, maptype = "toner-lite")
coords.map <- ggmap(coords.map, extent="device", legend="none")

coords.map.infecscale <- coords.map + stat_density2d(data=inatinfec,  aes(x=longitude, y=latitude, fill=..level.., alpha=..level..), geom="polygon")+scale_fill_gradientn(colours=rev(brewer.pal(7, "Spectral")))+ theme_bw()+xlab("Longitude")+ylab("Latitude")

coords.map+ geom_point(data=inatinfec,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")

coords.map.infecscale+ geom_point(data=inatinfec,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)
coords.map.infecscale

coords.map.scale <- coords.map + stat_density2d(data=inat,  aes(x=longitude, y=latitude, fill=..level.., alpha=..level..), geom="polygon")+scale_fill_gradientn(colours=rev(brewer.pal(7, "Spectral")))+ theme_bw()+xlab("Longitude")+ylab("Latitude")

coords.map+ geom_point(data=inat,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)+xlab("Longitude")+ylab("Latitude")

coords.map.scale+ geom_point(data=inat,  aes(x=longitude, y=latitude), fill="red",shape=21, size=1,alpha=0.3)
coords.map.scale
```


