---
title: "trillium_phytoplasma"
author: "Julia Boyle"
date: "2023-05-27"
output: html_document
---
# Field data

## Frequency of infection and hemiptera
look at prevalence vs presence more? 

```{r}
library(tidyverse)
library(cowplot)
library(patchwork)
library(car)
library(GLMMadaptive)
library(pscl)

trillium<-read.csv("Trillium_Field_Data_Frequency.csv")#just 2023
trillium<-trillium %>% mutate(Total_hemiptera_trillium=Young_hemiptera+Health_hemiptera+Infec_hemiptera)
trilliumY<-read.csv("Trillium_Field_Data - Multiyear.csv")#includes 2018 and 2023
trilliumY
trilliumY$Year<-as.factor(trilliumY$Year)
trilliumY
trillium_long<-trilliumY %>% 
  pivot_longer(cols=8:13,
  names_to = c("Status", "Hemiptera_presence"),
  names_sep = "_",
  values_to = "count"
)
trilliumY<-trilliumY %>% mutate(Infection_presence=ifelse(Freq_infection_of_adults>0,"Yes","No"))

trillium_long<-trillium_long %>% mutate(Infection= ifelse(Status=="Infec", "1", "0"))
str(trillium_long)
trillium_long$Infection<-as.integer(trillium_long$Infection)
trillium_long$Year<-as.factor(trillium_long$Year)
trillium_long_adults<-trillium_long %>% 
  filter(Status!="Young")
trillium_long
```

```{r}
hist(log(trilliumY$Freq_infection_of_adults+1))
hist(trilliumY$Freq_infection_of_adults)
hist(trilliumY$Freq_hemiptera_total)
hist(trilliumY$Total_trilliums)
```
Zero-inflated poisson models
```{r}
#2018 and 2023
##with random effects
head(trilliumY)
str(trilliumY)
MM4<-mixed_model(Total_infec_trilliums~Total_adult_trilliums+Year, random = ~ 1 | Population, data = trilliumY, family = zi.poisson(), zi_fixed = ~ Total_adult_trilliums+Year, zi_random=~1 |Population,control=list(iter_EM = 0))
summary(MM4) #loglik -254.7688. count model: density p=0.069. zero model:	 higher density increases likelihood of finding 0s p=0.01, higher year decreases chance of finding 0s=0.0388
##without random effects
summary(zeroinfl(Total_infec_trilliums~Total_adult_trilliums+Year, data=trilliumY)) #loglik -272.1, basically same results as above

#2023 only
##with random effects
head(trillium)
MM2<-mixed_model(Total_infec_trilliums~Freq_hemiptera_adults+Total_adult_trilliums, random = ~ 1 | Population, data = trillium, family = zi.poisson(), zi_fixed = ~ Freq_hemiptera_adults+Total_adult_trilliums, zi_random=~1 |Population,control=list(iter_EM = 0))
summary(MM2) #loglik -83.28, freq hemip 0.057 in count model 

##without random effects
summary(zeroinfl(Total_infec_trilliums~Freq_hemiptera_adults+Total_adult_trilliums, data=trillium)) #log lik -82.6, sig more likely to get zero infection in high adult density , marginal hemip in count model

#correlations
cor(trillium$Total_trilliums, trillium$Freq_hemiptera_total) #-0.475946
cor(trillium$Freq_infection_of_adults, trillium$Freq_hemiptera_adults) #0.3341377
cor(trillium$Freq_infection_of_adults, trillium$Young_to_repro_ratio) #0.1034085
```

```{r}
hist(trillium$Freq_hemiptera_adults)
hist(log(trillium$Freq_infection_of_adults))
hist(trillium$Freq_hemiptera_total)
hist(log(trillium$Total_trilliums))
hist(log(trillium$Young_to_repro_ratio))

summary(lm(Freq_infection_of_adults~Freq_hemiptera_adults, data=trillium))

summary(lm(Freq_hemiptera_total~Total_trilliums, data=trillium))

summary(lm(log(Young_to_repro_ratio)~Freq_infection_of_adults, data=trillium))
```


```{r}
trillium %>% ggplot(aes(x=Total_trilliums, y=Total_infec_trilliums, color=Freq_hemiptera_adults))+geom_point()+theme_cowplot()+labs(x="Trillium (no.)", y="Symptomatic trillium (no.)")
trillium %>% ggplot(aes(x=Total_adult_trilliums, y=Total_infec_trilliums, color=Freq_hemiptera_adults))+geom_point()+theme_cowplot()+labs(x="Flowering trillium (no.)", y="Symptomatic trillium (no.)")+scale_color_continuous("Freq. of hemiptera on flowering trillium",low = "blue", high = "hotpink")
```
```{r}
trilliumY %>% 
  ggplot(aes(x=Population, y=Freq_infection_of_adults, colour=Population))+geom_boxplot()+facet_wrap(~Forest_side)

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Forest_side))+geom_point()+geom_smooth(method='gam', se=F)+facet_wrap(~Year)

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Forest_side))+geom_point()+facet_wrap(~Year)+theme_cowplot()

trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method='gam', se=F)+facet_wrap(~Year)+theme_cowplot()+xlab("Density of flowering trillium in 1m^2")+ylab("Frequency of infected adults")
trilliumY %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults, colour=Year))+geom_point()+geom_smooth(method='gam', se=F)


trilliumY %>% filter(Year=='2023') %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_infection_of_adults))+geom_point()+xlab("Density of adults in 1m^2")+ylab("Frequency of infected adults")+theme_cowplot()

trilliumY %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm")
```

```{r}
hist(trillium$Freq_hemiptera_adults)
hist(trillium$Freq_infection_of_adult)
trillium %>%
  ggplot(aes(x=Freq_hemiptera_adults, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm", formula=y~x+0)+xlab("Frequency of hemiptera on adults")+ylab("Frequency of infected adults")+theme_cowplot()#+ylim(0,0.23) #forcing it to pass through 0??
```
```{r}
trillium %>% 
  ggplot(aes(x=Total_adult_trilliums, y=Freq_hemiptera_adults))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Density of flowering trilliums")+ ylab("Frequency of hemiptera on flowering trilliums")
trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_hemiptera_total))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Density of all trilliums no./m^2")+ ylab("Frequency of hemiptera on all trilliums")
#in very dense patches, hemiptera cannot colonize all the plants, meaning many will remain aphid-free and disease free
#in low density patches, almost all plants may get hemiptera, and so the likelihood of seeing infection is heightened
#predator satiation? 
#additionally, there is evidence sparse plant spacing can increase aphid population density (Chung et al. 1980) https://krex.k-state.edu/bitstream/handle/2097/35613/rec123-chung-1980.pdf?sequence=1&isAllowed=y
#survivorship bias? fewer surviving 
#defensive plant compounds released with herb, more effective at high density
```

```{r}
trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm")
trillium %>% 
  ggplot(aes(x=Freq_infection_of_adults, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm")+theme_cowplot()+xlab("Infection frequency of adults")+ylab("Young:Reproductive ratio")
```
Figures

```{r}
fig2A<-trilliumY %>% group_by(Total_adult_trilliums, Infection_presence, Year) %>% tally() %>% ggplot(aes(Total_adult_trilliums, fill = Infection_presence, colour = Infection_presence)) +
  geom_density(alpha = 0.1)+facet_grid(~Year)+theme_cowplot()+xlab(expression('Density of flowering trillium (no./1 m'^2* ')'))+ylab("Density of sampled plots")+scale_fill_manual(name="Infection presence",values=c("grey", "darkgreen"))+scale_colour_manual(name="Infection presence",values=c("grey", "darkgreen"))+labs(subtitle = "Year p = 0.038, Trillium density p = 0.010")+ theme(legend.position = "inside",legend.position.inside = c(0.25, 0.8))

fig2B<-trilliumY %>% group_by(Total_adult_trilliums, Total_infec_trilliums, Year) %>% tally() %>% ggplot(aes((Total_adult_trilliums), y=Total_infec_trilliums)) +
  geom_point(alpha=0.5)+facet_grid(~Year)+theme_cowplot()+xlab(expression('Density of flowering trillium (no./1 m'^2* ')'))+ylab("Number of infected trillium")+labs(subtitle = "Year p = 0.706, Trillium density p = 0.069")

cor.test(trillium$Total_trilliums, trillium$Freq_hemiptera_total)
fig2C<-trillium %>% 
  ggplot(aes(x=Total_trilliums, y=Freq_hemiptera_total))+geom_point()+geom_smooth(method="lm", color="darkgreen")+theme_cowplot()+xlab(expression('Density of all trillium (no./1 m'^2* ')'))+ ylab("Proportion of all trillium with hemiptera")+labs(subtitle = "r = -0.476, p < 0.001")

cor.test(trillium$Freq_hemiptera_adults, trillium$Freq_infection_of_adults)
fig2D<-trillium %>%
  ggplot(aes(x=Freq_hemiptera_adults, y=Freq_infection_of_adults))+geom_point()+geom_smooth(method="lm", color="darkgreen")+xlab("Proportion of flowering trillium with hemiptera")+ylab("Proportion of flowering trillium infected")+theme_cowplot()+labs(subtitle = "r = 0.334, p = 0.006")

plot_grid(fig2A,fig2B,fig2C, fig2D, labels=c("A", "B", "C", "D"))  #10 by 10
```

```{r}
cor.test(trillium$Freq_infection_of_adults, trillium$Young_to_repro_ratio)
demography2023<-trillium %>% 
  ggplot(aes(x=Freq_infection_of_adults, y=Young_to_repro_ratio))+geom_point()+geom_smooth(method="lm", color="darkgreen")+theme_cowplot()+xlab("Proportion of flowering trillium infected")+ylab("3-leaved non-flowering:flowering ratio")+labs(subtitle = "r = 0.103, p = 0.412")+ggtitle("2023")
demography2023
```

## Trait data
```{r}
traits<-read.csv("Trillium_Field_Data - Leaf and Height Phenotypes.csv")
str(traits)
traits<-traits %>% mutate(Infection_chr=ifelse(Infection=="0", "Asymptomatic", "Symptomatic"))
```
```{r}
#Height
figheight<-traits %>% ggplot(aes(x=Infection_chr, y=Height_cm))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Height (cm)")
traits %>% ggplot(aes(x=Infection, y=Height_cm))+geom_point()+geom_smooth()+theme_cowplot()
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Height_cm))
1-(23.34688/27.61000	)
27.61-23.34688

#leaf length
traits %>% ggplot(aes(x=Infection_chr, y=Leaf_length_cm))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Leaf length (cm)")

#sepal length
figsepal<-traits %>% ggplot(aes(x=Infection_chr, y=Sepal_length_cm))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Sepal length (cm)")
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Sepal_length_cm))
1-(3.460667/4.820938)

#leaf and sepal ratio
figratio<-traits %>% ggplot(aes(x=Infection_chr, y=Sepal_to_leaf_ratio))+geom_boxplot()+theme_cowplot()+labs(x = "Infection status",y = "Sepal:Leaf length ratio")
traits %>%  group_by(Infection_chr) %>% summarize(mean=mean(Sepal_to_leaf_ratio))

#figure
plot_grid(figheight,figsepal, figratio, labels=c("A", "B", "C"))
```
```{r}
hist(traits$Height_cm)
summary(lmer(Height_cm~Infection_chr+(1|Plot/Population), data=traits))
Anova(lmer(Height_cm~Infection_chr+(1|Plot/Population), data=traits), type=3) #p=0.003096

hist(traits$Leaf_length_cm)
Anova(lmer(Leaf_length_cm~Infection_chr+(1|Plot/Population), data=traits), type=3) #not sig

hist(log(traits$Sepal_length_cm))
shapiro.test(log(traits$Sepal_length_cm))
summary(lmer(log(Sepal_length_cm)~Infection_chr+(1|Plot/Population), data=traits))
Anova(lmer(log(Sepal_length_cm)~Infection_chr+(1|Plot/Population), data=traits), type=3) #p=2.342e-07
```

## Sticky trap data
```{r}
sticky<-read.csv("Trillium_Field_Data - Sticky traps.csv")
head(sticky)
str(sticky)
sticky$Adult_Trillium_density<-factor(sticky$Adult_Trillium_density, levels=c("Low", "Mid", "High"))
sticky_noout<-sticky %>% filter(Outlier=="No")
```

```{r}
sticky_noout %>% ggplot(aes(x=Adult_Trillium_density, y=Num_sapsuckers_on_trap))+geom_boxplot()
sticky_noout %>% ggplot(aes(x=Adult_Trillium_density, y=Total_insects_on_trap))+geom_boxplot()
sticky_noout %>% ggplot(aes(x=Forest, y=Num_sapsuckers_on_trap, fill=Adult_Trillium_density))+geom_boxplot()+theme_cowplot()+scale_fill_grey(name="Density of flowering trillium", start=0.5, end=1)+labs(x = "Forest population",y = "Hemiptera caught (no.)")+scale_x_discrete(labels=c("Old growth", "Secondary"))
sticky_noout %>% ggplot(aes(x=Forest, y=Total_insects_on_trap, color=Adult_Trillium_density))+geom_boxplot()
#this mainly shows that old forest had more bugs than secondary forest, and that intermediate trillium density plots tended to have the most insects

sticky_noout %>% ggplot(aes(x=Num_sapsuckers_on_trap, y=Freq_infection_of_adults, color=Adult_Trillium_density))+geom_point()+theme_cowplot()
sticky_noout %>% ggplot(aes(x=Num_sapsuckers_on_trap, y=Freq_hemiptera_total, color=Adult_Trillium_density))+geom_point()+theme_cowplot()
```

## Seed counts 2024
```{r}
#actual seed counts only
seed<-read.csv("2024 Trillium seed counts and predicted counts.csv")
head(seed)
seed$seed_count_round<-round(seed$seed_count)
seed<-seed %>% mutate(Full_treatment=paste(Infection_status, Seed_count_real_or_predicted, sep = '_'))

seed %>% drop_na(seed_count) %>%  ggplot(aes(x=Infection_status, y=seed_count, colour=Seed_count_real_or_predicted))+geom_boxplot()+geom_point( position=position_jitterdodge(jitter.width=0.25, dodge.width=0.8), alpha=0.2)+theme_cowplot()+ylab("Seed count")+xlab("Infection status")+scale_x_discrete(labels=c("Asymptomatic", "Symptomatic"))+scale_colour_manual(values=c("goldenrod", "darkgrey"),name="Seed count", labels=c("Predicted from leaf length", "Real"))+theme(legend.position="top")

seed_tril_plot<-seed %>% drop_na(seed_count) %>%filter(Full_treatment!="Nonsymptomatic_Real") %>%   ggplot(aes(x=Infection_status, y=seed_count, colour=Seed_count_real_or_predicted))+geom_boxplot()+geom_point( position=position_jitterdodge(jitter.width=0.25, dodge.width=0.8), alpha=0.2)+theme_cowplot()+ylab("Seed count")+xlab("Infection status")+scale_x_discrete(labels=c("Asymptomatic", "Symptomatic"))+scale_colour_manual(values=c("black", "darkgreen"),name="Seed count", labels=c("Predicted from leaf length", "Real"))+theme(legend.position="top")

seed %>%filter(Paired_data_plants=='Yes'& Seed_count_real_or_predicted=='Real') %>%  drop_na(seed_count)%>% group_by(Infection_status) %>% tally() #only 81 plants had real seed count numbers
seed %>%filter(Paired_data_plants=='Yes') %>% group_by(Infection_status) %>% filter(seed_count==0) %>% tally()
51/81
seed_comparison1<-seed %>% filter(Paired_data_plants=='Yes'&Infection_status=="Symptomatic")
seed_comparison2<-seed %>% filter(Full_treatment=="Symptomatic_Real" | Full_treatment== "Nonsymptomatic_Predicted_from_lf_length")
hist((seed_comparison2$seed_count))
Anova(glmer.nb(round(seed_count)~Seed_count_real_or_predicted+(1|pair_number), data=seed_comparison1), type=3) #comparing infected observed vs. expected seed number based on size
(Anova(glmer.nb(round(seed_count)~Infection_status+(1|pair_number), data=seed_comparison2, control = glmerControl("bobyqa")), type=3)) #comparing the focal paired plants, actual infected and predicted seed count of healthy paired plant

#try quasipoisson if overdispersed
plyr::ddply(seed_comparison2, c("Infection_status"), summarise,
                  N=sum(!is.na(seed_count)),
             mean = mean ((seed_count), na.rm=TRUE),
             sd = sd (seed_count,na.rm=TRUE),
             se = sd/sqrt(N))
((5.851852		- 18.525928	)/18.525928)*100
plyr::ddply(seed_comparison1, c("Seed_count_real_or_predicted"), summarise,
                  N=sum(!is.na(seed_count)),
             mean = mean ((seed_count), na.rm=TRUE),
             sd = sd (seed_count,na.rm=TRUE),
             se = sd/sqrt(N))
((5.851852		- 16.522099	)/16.522099)*100
```

```{r}
plot_grid(seed_tril_plot, demography2023, labels=c("A", "B")) #5 by 8 landscape for ellie's version
#5 by 10 for the above version

plot_grid(seed_tril_plot, demography2024_one, labels=c("A", "B"))

#if use demography plots from 2024 data, make them in other trillium code doc
plot_grid(seed_tril_plot, demography2023, demography2024_one, demography_2024_veg, labels="AUTO") #8 by 10
```


# iNaturalist data
```{r}
#install.packages('geodata')
library(maps)
library(mapdata)
library(ggmap)
library(RColorBrewer)
library(raster)
library(scales)
library(CoordinateCleaner)
library(rnaturalearth)
library(rnaturalearthdata)
library(geodata)
library(sf)
library(plyr)
library(spData)
library(tidyterra)
library(terra)
library(tmap) 
library(predicts)
library(tidyverse)
```

```{r}
#inat<-read.csv("iNat_alltrilliums_2023_07_05.csv") #only up until 2023
inat<-read.csv("iNat_alltrilliums_2025_03_14_539340.csv") #version with 2024 data
head(inat)

inat_cleaned = inat %>% 
  # Adding positional accuracy filter to see if that helps deal with observations in lakes
  filter(!is.na(latitude), !is.na(longitude), !is.na(public_positional_accuracy)) %>% 
  filter(public_positional_accuracy < 10000) %>% 
  dplyr::mutate(infected = `field.phytoplasma.infection.`) %>% 
  mutate(infected = if_else(is.na(infected), "No", infected)) %>% 
  # Drop european and puerto rico observations
  filter(latitude > 20, longitude < -65) %>% 
  # Group by user and subset to 15 observations per user
  group_by(user_login) %>% 
  slice_sample(n = 15, replace = FALSE)

inat_clean <- clean_coordinates(x = inat_cleaned, 
                           lon = "longitude", 
                           lat = "latitude",
                           species = "scientific_name",
                           tests = c("centroids",
                                    "equal", "zeros","institutions", "duplicates")) 
inat_clean=inat_clean %>% filter(.summary==TRUE) #only include observations that have no suspect issues
inat_clean=inat_clean %>% mutate(infected=ifelse(grepl("Yes", infected), "1", "0"))
inat_clean$infected

inat%>% filter( field.phytoplasma.infection.=="Yes") #845
inat_cleaned%>% filter( field.phytoplasma.infection.=="Yes") #550
inat_clean%>% filter( field.phytoplasma.infection.=="Yes") #526 after filtering out stuff with clean_coordinates
```
```{r}
#old code to do the above cleaning of users
set.seed(1)
overrepresented_users<-inat_clean %>% group_by(user_login) %>% tally() %>% filter(n>15)
overrepresented_samples<-subset(inat_clean, user_login %in% overrepresented_users$user_login)
overrepresented_subset<-plyr::ddply(overrepresented_samples,.(user_login),function(x) x[sample(nrow(x),15),])#make the max number of observations per user 15 
normalrepresented_users<-inat_clean %>% group_by(user_login) %>% tally() %>% filter(n<=15)
normalrepresented_samples<-subset(inat_clean, user_login %in% normalrepresented_users$user_login)
inat_clean_corrected<-rbind(overrepresented_subset,normalrepresented_samples)
inat_clean_corrected

inat_clean_tally<-inat_clean %>% group_by(user_login) %>% tally()
mean(inat_clean_tally$n)
median(inat_clean_tally$n)

inatinfec<-inat_clean %>% filter( field.phytoplasma.infection.=="Yes") #make another dataset of infected points
inatinfec %>% group_by(user_login) %>% tally() #vast majority of users contributed less than 6 infected observations, one person contributed 12
inat_clean %>% group_by(user_login) %>% tally()

inat_clean %>% filter(observed_on>'2020-01-01') #23405
23405/28390
```

## Extracting coordinate data
Run this everytime you load R because spatrasters do not save with R enviro
```{r}
dir.create(path = "data") 
dir.create(path = "output")
bioclim_data_source <- worldclim_global(var = "bio", res = 0.5, path = "data/")
#Determine geographi cextent of our data 
max_lat<-ceiling(max(inat_clean$latitude)) 
min_lat<-floor(min(inat_clean$latitude)) 
max_lon<-ceiling(max(inat_clean$longitude)) 
min_lon<-floor(min(inat_clean$longitude)) 
#Store boundaries in a single extent object 
geographic_extent<-ext(x=c(min_lon,max_lon,min_lat,max_lat))

# Make an extent that is 25% larger 
sample_extent <- geographic_extent * 1.25 
# Crop bioclim data to desired extent 
bioclim_data <- crop(x = bioclim_data_source, y = sample_extent) 
# Plot the first of the bioclim variables to check on cropping 
plot(bioclim_data[[1]])
plot(bioclim_data[[12]])

bioclim_extract <- terra::extract(x = bioclim_data, y = as.matrix(inat_clean[, c("longitude", "latitude")])) # No need for an ID column
# I GET 41 OBSERVATIONS WITH NAS FROM BIOCLIM DATA- WHY???
# Add the point and climate datasets together 
points_climate <- cbind(inat_clean, bioclim_extract)

crop_data<-cropland(source="WorldCover",  path = "data/")
crop_data<- crop(x = crop_data, y = sample_extent) 
plot(crop_data[[1]])
cropdata_extract<- terra::extract(x = crop_data, y = as.matrix(inat_clean[, c("longitude", "latitude")]))
points <- cbind(points_climate, cropdata_extract)

#stonefruit<-crop_monfreda(crop="stonefruitnes", var="area_ha",  path = "data/") #download failed?

footprint<-footprint(year=2009, path = "data/")
footprint_crop<- crop(x = footprint, y = sample_extent) 
plot(footprint_crop[[1]])
footprint_extract<- terra::extract(x = footprint_crop, y = as.matrix(inat_clean[, c("longitude", "latitude")]))
footprint_extract<-rename(footprint_extract, footprint = `wildareas-v3-2009-human-footprint`
)
points<-cbind(points, footprint_extract)

bioclim_data
footprint_crop
crop_data
geo_data<-c(bioclim_data, footprint_crop, crop_data)
geo_data_test<-geo_data
```

##Density map
```{r}
dat_sf = st_as_sf(inat_clean, coords = c("longitude", "latitude")) %>% 
  st_set_crs("EPSG:4326")
# Reproject to equal area projection
dat_ea = st_transform(dat_sf, crs = "ESRI:102003")
# Make a grid of polygons that encompass the observation points
# Choosing an arbitrary 5 km grid size
area_fishnet_grid = st_make_grid(dat_ea, c(5e4, 5e4), what = "polygons", square = FALSE)

# Convert to sf and add grid ID
fishnet_grid_sf = st_sf(area_fishnet_grid) %>%
  # add grid ID
  mutate(grid_id = 1:length(lengths(area_fishnet_grid)))
# Assign observations to grid cells
grid_assignments = st_intersection(fishnet_grid_sf, dat_ea)

# Calculate frequencies per grid cell
freq_per_poly = grid_assignments %>% 
  group_by(grid_id) %>% 
  summarize(n = n(), n_infected = sum(infected == "Yes")) %>% 
  mutate(freq_infected = n_infected/n)

# Rejoin these calculated frequencies to polygons for plotting
freqs_with_polys = st_join(fishnet_grid_sf, freq_per_poly) %>% 
  filter(!is.na(n))

world = ne_countries(scale = "large", returnclass = "sf", continent = "north america")
lakes = ne_download(scale = "medium", type = "lakes", category = "physical")

tril_b<-ggplot() +
  geom_sf(data = world, fill = "white") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  geom_sf(data = freqs_with_polys, aes(fill = freq_infected), alpha = 0.8, color = "lightgrey") +
  scale_fill_gradientn(colours = c("white", "#CCFF99", "darkgreen"), values = c(0,0.000001,1), name="Proportion \ninfected") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()

tril_a<-ggplot() +
  geom_sf(data = world, fill = "white") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  geom_sf(data = freqs_with_polys, aes(fill = n), alpha = 0.8, color = "lightgrey") +
  scale_fill_gradientn(colours = c("white", "lightblue", "navy"), name="Observations") +
  coord_sf(xlim = c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()

plot_grid(tril_a, tril_b, labels=c("A", "B"), ncol=1) #10 by 7 portrait
trilab<-plot_grid(tril_a, tril_b, labels=c("A", "B"), ncol=1)
```
##Moran’s I analysis
https://mgimond.github.io/simple_moransI_example/
```{r}
library(sf)
library(spdep)
library(tmap)
nb <- poly2nb(freqs_with_polys, queen=TRUE)
lw <- nb2listw(nb, style="W", zero.policy=TRUE)
inc.lag <- lag.listw(lw, freqs_with_polys$n_infected)
plot(inc.lag ~ freqs_with_polys$n_infected, pch=16, asp=1)
M1 <- lm(inc.lag ~ freqs_with_polys$n_infected)
abline(M1, col="blue")
coef(M1)[2] #0.4216945
#this slope is Morans I: because this is positive, it means polygons with high frequency of infection are more likely to have neighbouring polygons with high frequency of infection
#a positive Moran’s I value suggests clustering whereas a negative Moran’s I value suggests dispersion
inc.lag.freq <- lag.listw(lw, freqs_with_polys$freq_infected)
plot(inc.lag.freq ~ freqs_with_polys$freq_infected, pch=16, asp=1)
M1.freq <- lm(inc.lag.freq ~ freqs_with_polys$freq_infected)
abline(M1.freq, col="blue")
coef(M1.freq)[2] #0.1072854

moran(freqs_with_polys$n_infected, lw, length(nb), Szero(lw))[1] #0.4302413, infections are pretty clustered in space
moran(freqs_with_polys$freq_infected, lw, length(nb), Szero(lw))[1] #0.1082928
moran(freqs_with_polys$n, lw, length(nb), Szero(lw))[1] #0.3096669

MC.ninfec<- moran.mc(freqs_with_polys$n_infected, lw, nsim=999, alternative="greater")
MC.ninfec
plot(MC.ninfec)
MC.freq<- moran.mc(freqs_with_polys$freq_infected, lw, nsim=999, alternative="greater")
MC.freq
plot(MC.freq)
MC.n<- moran.mc(freqs_with_polys$n, lw, nsim=999, alternative="greater")
MC.n
plot(MC.n)
```

##Models
what seems to cause infection?
```{r}
as.data.frame(grid_assignments)
as.data.frame(freq_per_poly)

inat_grid<-merge(as.data.frame(grid_assignments), as.data.frame(freq_per_poly), by.x="grid_id", by.y="grid_id")
inat_grid<-inat_grid %>% dplyr::select(grid_id, id, n, n_infected, freq_infected)

inat_full<-merge(inat_grid, points, by.x="id", by.y="id")
inat_full$infected<-as.numeric(inat_full$infected)
inat_full$footprint<-inat_full$`wildareas-v3-2009-human-footprint`
inat_full$Year<-substring(inat_full$observed_on,1,4)
str(inat_full)
```
```{r}
#add total trillium number as a layer to the raster 
r <- rast(xmin=-100.875, xmax=-62.125, ymin=31.125, ymax=49.875, ncols=4650, nrows=2250, crs = crs(geo_data_test))
# you need to provide a function 'fun' for when there are multiple points per cell
trillium_number <- rasterize(as.matrix((inat_full %>% select(longitude, latitude,n))[, 1:2]), r, as.matrix((inat_full %>% select(longitude, latitude,n))[,3]), fun=sum)
plot(trillium_number_resampled)
trillium_number_resampled <- resample(trillium_number, geo_data_test)
names(trillium_number_resampled)<-"n"
geo_data_test<-c(geo_data_test, trillium_number_resampled)

#add lat and long as a layer to the raster??
latitude <- rasterize(as.matrix((inat_full %>% select(longitude, latitude))[, 1:2]), r, as.matrix(inat_full$latitude), fun=mean)
latitude<- resample(latitude, geo_data_test)
names(latitude)<-"latitude"
geo_data_test<-c(geo_data_test, latitude)
longitude <- rasterize(as.matrix((inat_full %>% select(longitude, latitude))[, 1:2]), r, as.matrix(inat_full$longitude), fun=mean)
names(longitude)<-"longitude"
longitude<- resample(longitude, geo_data_test)
geo_data_test<-c(geo_data_test, longitude)
plot(longitude)
```

###GAM
determine which variables to keep in model (reduce collinearity)
```{r}
library(usdm)
s(n)+s(wc2.1_30s_bio_1)+ s(wc2.1_30s_bio_2)+s(wc2.1_30s_bio_3)+s(wc2.1_30s_bio_4)+s(wc2.1_30s_bio_5)+s(wc2.1_30s_bio_6)+s(wc2.1_30s_bio_7)+s(wc2.1_30s_bio_8)+s(wc2.1_30s_bio_9)+s(wc2.1_30s_bio_10)+s(wc2.1_30s_bio_11)+s(wc2.1_30s_bio_12)+s(wc2.1_30s_bio_13)+s(wc2.1_30s_bio_14)+s(wc2.1_30s_bio_15)+s(wc2.1_30s_bio_16)+s(wc2.1_30s_bio_17)+s(wc2.1_30s_bio_18)+s(wc2.1_30s_bio_19)+s(cropland)+s(footprint)

vifcor(inat_full[53:73], th=0.9)
#this would exclude wc2.1_30s_bio_17, wc2.1_30s_bio_6, wc2.1_30s_bio_16, wc2.1_30s_bio_11, wc2.1_30s_bio_10, wc2.1_30s_bio_19, wc2.1_30s_bio_7, wc2.1_30s_bio_12

s(n, k=50)+s(wc2.1_30s_bio_1)+ s(wc2.1_30s_bio_2)+s(wc2.1_30s_bio_3)+s(wc2.1_30s_bio_4)+s(wc2.1_30s_bio_5)+s(wc2.1_30s_bio_8)+s(wc2.1_30s_bio_9)+s(wc2.1_30s_bio_13)+s(wc2.1_30s_bio_14)+s(wc2.1_30s_bio_15)+s(wc2.1_30s_bio_18)+s(cropland)+s(footprint)+s(latitude,longitude,bs="gp",m=2,k=200)

vifstep(inat_full[53:73])
#this would exclude wc2.1_30s_bio_6, wc2.1_30s_bio_11, wc2.1_30s_bio_1, wc2.1_30s_bio_3, wc2.1_30s_bio_4, wc2.1_30s_bio_17, wc2.1_30s_bio_12, wc2.1_30s_bio_5, wc2.1_30s_bio_16, wc2.1_30s_bio_19, wc2.1_30s_bio_14, wc2.1_30s_bio_13 

s(n, k=50)+ s(wc2.1_30s_bio_2)+s(wc2.1_30s_bio_7)+s(wc2.1_30s_bio_8)+s(wc2.1_30s_bio_9)+s(wc2.1_30s_bio_10)+s(wc2.1_30s_bio_15)+s(wc2.1_30s_bio_18)+s(cropland)+s(footprint)+s(latitude,longitude,bs="gp",m=2,k=200)
```

Run models
```{r}
library(mgcv)
#https://m-clark.github.io/generalized-additive-models/appendix.html
#https://r.qcbs.ca/workshop08/book-en/gam-model-checking.html
#https://noamross.github.io/gams-in-r-course/     <-----THIS ONE IS GREAT!!!
# +s(latitude,longitude,bs="sos",m=2,k=200) #Spherical version
#+s(latitude,longitude,bs="gp",m=2,k=200) #gaussian version
gam_tril_allsmooth<-gam(infected ~ s(n, k=50) +s(wc2.1_30s_bio_1) +s(wc2.1_30s_bio_4) +s(wc2.1_30s_bio_5) +s(wc2.1_30s_bio_6) +s(wc2.1_30s_bio_12) +s(wc2.1_30s_bio_13) +s(wc2.1_30s_bio_14) +s(wc2.1_30s_bio_15) +s(cropland) +s(footprint) +s(latitude,longitude,bs="gp",m=2,k=200),family="binomial",method = "REML",data=inat_full)

#model 1
summary(gam_tril_allsmooth)
k.check(gam_tril_allsmooth) #if edf is too close to k, gives sig value. we want k-index near 1 and nonsig p. k controls wiggliness
set.seed(1)
gam.check(gam_tril_allsmooth)
plot(gam_tril_allsmooth, residuals = TRUE, pch = 1, all.terms = TRUE, shade = TRUE, shift = coef(gam_tril)[1])
concurvity(gam_tril_allsmooth, full = TRUE) #if worst case has values over 0.8 then you have problem similar to colinearity but with parabola-ness. Use full=F to find out the offenders
concurvity(gam_tril_allsmooth, full = F)
rpred <- predict(geo_data, gam_tril_allsmooth)#but maybe add one layer with tril n. eg. rstack <- stack(rx1, rx2, rzeros)/ names(rstack) <- c("x1", "x2", "sitef")
vis.gam(gam_tril_allsmooth, view = c("latitude", "longitude"), plot.type = "contour", too.far = 0.05)#means it won't extrapolate more than 5% into untested combinations

#model 2
set.seed(1)
gam_vifstep<-gam(infected ~ s(n, k=50)+ s(wc2.1_30s_bio_2)+s(wc2.1_30s_bio_7)+s(wc2.1_30s_bio_8)+s(wc2.1_30s_bio_9)+s(wc2.1_30s_bio_10)+s(wc2.1_30s_bio_15)+s(wc2.1_30s_bio_18, k=30)+s(cropland)+s(footprint)+s(latitude,longitude,bs="gp",m=2,k=200),family="binomial",method = "REML",data=inat_full)
summary(gam_vifstep)
gam.check(gam_vifstep)
vis.gam(gam_vifstep, view = c("longitude", "latitude"), plot.type = "contour", too.far = 0.05, color="topo")#means it won't extrapolate more than 5% into untested combinations
plot(gam_vifstep, residuals = TRUE, pch = 1, all.terms = TRUE, shade = TRUE, shift = coef(gam_tril)[1])
```

Species distribution model?
https://jcoliver.github.io/learn-r/011-species-distribution-models.pdf 
```{r}
inat_1<-merge(inat_grid, points, by.x="id", by.y="id")
inat_1 = subset(inat_1, select = -c(.val,.equ, .zer, .cen, .inst, .dpl, .summary) )
# Create vector indicating fold for making a training dataset 
fold <- folds(x = inat_1, k = 5, by = inat_full$infected) #five folds so that we can test on 20% of the data
testing <- inat_1[fold == 1, ] 
training <- inat_1[fold != 1, ]
training$footprint<-as.numeric(training$footprint)
training$infected<-as.numeric(training$infected)
str(training)
# Build a model using training data 
gam_training<-gam(infected ~ s(n, k=50)+ s(wc2.1_30s_bio_2)+s(wc2.1_30s_bio_7)+s(wc2.1_30s_bio_8)+s(wc2.1_30s_bio_9)+s(wc2.1_30s_bio_10)+s(wc2.1_30s_bio_15)+s(wc2.1_30s_bio_18, k=30)+s(cropland)+s(footprint)+s(latitude,longitude,bs="gp",m=2,k=200),family="binomial",method = "REML",data=training)
summary(gam_training)

which(is.na(training$footprint))
names(geo_data_test)
# Get predicted values from the model 
gam_predict <- predict(geo_data_test, gam_training, type = "response") #geo_data but with trillium number and lat and long
# Print predicted values 
plot(gam_predict)
# Use testing data for model evaluation 
gam_eval <- pa_evaluate(p = testing[testing$infected == 1, ], a = testing[testing$infected == 0, ], model = gam_training, type = "response")
# Determine minimum threshold for "presence" 
gam_threshold <- gam_eval@thresholds$max_spec_sens
```
```{r}
world = ne_countries(scale = "large", returnclass = "sf", continent = "north america")
lakes = ne_download(scale = "medium", type = "lakes", category = "physical")

tril_c<-ggplot() +
   geom_spatraster(data = (gam_predict> gam_threshold), aes(fill = lyr1))+#Only plot areas where probability of occurrence is greater than the threshold 
  geom_sf(data = world, fill="NA") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  coord_sf(xlim =c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()+scale_fill_manual(name="Suitability for phytoplasma", values=c("NA","red"), na.value="NA")+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), shape=3, size=1,alpha=0.3)
#the map we have drawn presents a categorical classification of whether a particular point on the landscape will be suitable or not for the species of interest.
```


### INLA
https://ourcodingclub.github.io/tutorials/inla/
```{r}
#if(!require(ggregplot)) devtools::install_github("gfalbery/ggregplot") # Installing Greg's package for plotting functions!
#install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA); library(ggregplot); library(RColorBrewer)
head(inat_full)

phen <- c("longitude", "latitude") # Base columns with spatial information we'll need

resp <- "infected" # Response variable

covar <- c("n","wc2.1_30s_bio_1",  "wc2.1_30s_bio_2","wc2.1_30s_bio_3","wc2.1_30s_bio_4","wc2.1_30s_bio_5","wc2.1_30s_bio_6","wc2.1_30s_bio_7","wc2.1_30s_bio_8","wc2.1_30s_bio_9","wc2.1_30s_bio_10","wc2.1_30s_bio_11","wc2.1_30s_bio_12","wc2.1_30s_bio_13","wc2.1_30s_bio_14","wc2.1_30s_bio_15","wc2.1_30s_bio_16","wc2.1_30s_bio_17","wc2.1_30s_bio_18","wc2.1_30s_bio_19","cropland", "footprint")

# Specify the formula
f0.1 <- as.formula(paste0(resp, " ~ ", # Response first
                          paste(covar, collapse = " + ") # Collapse the vector of covariates
))

# Run the model 
IM0.1  <- inla(f0.1, 
               family = "binomial", # Specify the family. Can be a wide range (see r-inla.org).
               data = inat_full) # Specify the data
summary(IM0.1)
Efxplot(list(IM0.1)) #There are no P values in INLA. Importance or significance of variables can be deduced by examining the overlap of their 2.5% and 97.5% posterior estimates with zero.

#Perform model selection by removing covariates one by one and seeing how this changes model fit according to the model’s Deviance Information Criterion
#library(beepr)
#library(MCMCglmm)
ModelSel <- INLAModelSel(resp, covar, Family="binomial", Data=inat_full)
#[1] "Losing wc2.1_30s_bio_16" "Run 2" "Losing wc2.1_30s_bio_1" "Run 3""Losing wc2.1_30s_bio_18" "Run 4" "Losing wc2.1_30s_bio_17" "Run 5" "Losing wc2.1_30s_bio_13" "Run 6""Losing wc2.1_30s_bio_11" "Run 7" "Losing n" "Run 8" "Losing wc2.1_30s_bio_5" "Run 9" "Losing wc2.1_30s_bio_15" "Run 10" #######it removed 9, but I want to keep n in the model
Finalcovar<- c("n","wc2.1_30s_bio_2","wc2.1_30s_bio_3","wc2.1_30s_bio_4","wc2.1_30s_bio_6","wc2.1_30s_bio_7","wc2.1_30s_bio_8","wc2.1_30s_bio_9","wc2.1_30s_bio_10","wc2.1_30s_bio_12","wc2.1_30s_bio_14","wc2.1_30s_bio_19","cropland", "footprint")
#Finalcovar <- ModelSel$Removed[[length(ModelSel$Removed)]]
f1 <- as.formula(paste0(resp, " ~ ", 
                        paste(Finalcovar, collapse = " + "))) 

IM1 <- inla(f1,
            family = "binomial",
            data = inat_full,
            control.compute = list(dic = TRUE)) 
summary(IM1)
Efxplot(list(IM1))
##########################################################################
which(is.na(inat_full$wc2.1_30s_bio_1)) #41 uninfected trillium near quebec and montreal don't have associated bioclim data for some reason. we need to exclude them for INLA
sum(is.na(inat_full$wc2.1_30s_bio_1))
inat_full[which(is.na(inat_full$wc2.1_30s_bio_1)),]
inat_full_complete = inat_full[complete.cases(inat_full[53:73]),]
##########################################################################
#incorporating space
Locations = cbind(inat_full_complete$longitude, inat_full_complete$latitude) # using the sampling locations
MeshB <- inla.mesh.2d(Locations, max.edge = c(20, 40)) #bigger triangles for data exploration
MeshC <- inla.mesh.2d(Locations, max.edge = c(10, 20)) #smaller triangles- use smaller triangles for publication
Mesh <- MeshB
plot(MeshB)
plot(MeshC)
points(Locations, col = "red", pch = 2)
# Making the A matrix
HostsA <- inla.spde.make.A(Mesh, loc = Locations) # Making A matrix, which essentially translates spatial locations on the mesh into vectors in the model.
Hosts.spde = inla.spde2.pcmatern(mesh = Mesh, prior.range = c(10, 0.5), prior.sigma = c(.5, .5)) # Making SPDE
w.Host <- inla.spde.make.index('w', n.spde = Hosts.spde$n.spde) # making the w

# Making the model matrix #### 
X0 <- model.matrix(as.formula(paste0("~ -1 +", paste(Finalcovar, collapse = "+"))), data = inat_full_complete) # make the model matrix using the final model selection formula without a response variable.
X <- as.data.frame(X0)
# Making the stack ####
N <- nrow(inat_full_complete)
StackHost <- inla.stack(
  data = list(y = inat_full_complete[,resp]), # specify the response variable
  A = list(1,1,HostsA), # Vector of Multiplication factors for random and fixed effects  #put as many 1s as there are fixed and random effects in the effects() below            
  effects = list(
    Intercept = rep(1, N), # specify the manual intercept!
    X = X, # attach the model matrix
    # insert vectors of any random effects (none here)
    w = w.Host)) # attach the w 
#Running the model
f1 <- as.formula(paste0("y ~ -1 + Intercept + ", paste0(colnames(X), collapse = " + "))) #version only fixed effects
f3 <- as.formula(paste0("y ~ -1 + Intercept + ", paste0(colnames(X), collapse = " + "), " + f(w, model = Hosts.spde)")) #version with space SPDE random effect
IM1 <- inla(f1, # Base model (no random effects)
            family = "binomial",
            data = inla.stack.data(StackHost),
            control.compute = list(dic = TRUE),
            control.predictor = list(A = inla.stack.A(StackHost)))
IM3 <- inla(f3, # SPDE random effect 
            family = "binomial",
            data = inla.stack.data(StackHost),
            control.compute = list(dic = TRUE),
            control.predictor = list(A = inla.stack.A(StackHost)))
SpatialHostList <- list(IM1, IM3)

ggField(IM3, Mesh, Groups = 1) +
  scale_fill_brewer(palette = "Blues") #Plotting the spatial field 
sapply(SpatialHostList, function(f) f$dic$dic) #does space make it better?
INLADICFig(SpatialHostList, ModelNames = c("Base", "SPDE")) #if DIC is smaller by >2, then suggests autocorrelation was important to include
```


###Species distribution model?
https://jcoliver.github.io/learn-r/011-species-distribution-models.pdf 
```{r}
inat_1<-merge(inat_grid, points, by.x="id", by.y="id")
inat_1 = subset(inat_1, select = -c(.val,.equ, .zer, .cen, .inst, .dpl, .summary) )
# Create vector indicating fold for making a training dataset 
fold <- folds(x = inat_1, k = 5, by = inat_full$infected) #five folds so that we can test on 20% of the data
testing <- inat_1[fold == 1, ] 
training <- inat_1[fold != 1, ]

# Build a model using training data 
glm_model <- glm(as.numeric(infected) ~ wc2.1_30s_bio_1+ wc2.1_30s_bio_2+wc2.1_30s_bio_3+wc2.1_30s_bio_4+wc2.1_30s_bio_5+wc2.1_30s_bio_6+wc2.1_30s_bio_7+wc2.1_30s_bio_8+wc2.1_30s_bio_9+wc2.1_30s_bio_10+wc2.1_30s_bio_11+wc2.1_30s_bio_12+wc2.1_30s_bio_13+wc2.1_30s_bio_14+wc2.1_30s_bio_15+wc2.1_30s_bio_16+wc2.1_30s_bio_17+wc2.1_30s_bio_18+wc2.1_30s_bio_19+cropland+`wildareas-v3-2009-human-footprint`, data = training, family = binomial)
summary(glm_model)

tril_num<-rast(nrows=inat_full$latitude, ncols=inat_full$longitude, vals=inat_full$n)
geo_data_test<-geo_data
add(geo_data_test) <- tril_num

# Get predicted values from the model 
glm_predict <- predict(geo_data, glm_model, type = "response") #should I incorporate trillium n into this?
# Print predicted values 
plot(glm_predict)
# Use testing data for model evaluation 
glm_eval <- pa_evaluate(p = testing[testing$infected == 1, ], a = testing[testing$infected == 0, ], model = glm_model, type = "response")
# Determine minimum threshold for "presence" 
glm_threshold <- glm_eval@thresholds$max_spec_sens
```
```{r}
world = ne_countries(scale = "large", returnclass = "sf", continent = "north america")
lakes = ne_download(scale = "medium", type = "lakes", category = "physical")

tril_c<-ggplot() +
   geom_spatraster(data = (glm_predict> glm_threshold), aes(fill = lyr1))+#Only plot areas where probability of occurrence is greater than the threshold 
  geom_sf(data = world, fill="NA") +
  geom_sf(data = lakes, color = "grey", fill = "grey") +
  coord_sf(xlim =c(-102.15, -68), ylim = c(30, 55), expand = FALSE) +
  theme_bw()+scale_fill_manual(name="Suitability for phytoplasma", values=c("NA","red"), na.value="NA")+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), shape=3, size=1,alpha=0.3)
#the map we have drawn presents a categorical classification of whether a particular point on the landscape will be suitable or not for the species of interest.
```

```{r}
plot_grid(trilab, tril_c, labels=c("","C"), ncol=2)
```


##############################################################################################
##Old Main maps
```{r}
#http://mbontrager.org/blog/2016/08/01/Maps-in-R
infectedpoints <- filter(inat_clean_corrected, field.phytoplasma.infection.=="Yes")
normalpoints <- filter(inat_clean_corrected, field.phytoplasma.infection.=="")
infcolour <- brewer.pal(6, "Paired")[5]
normcolour <- brewer.pal(6, "Paired")[3]

usa <- map_data("state")
canada <- map_data("worldHires","Canada")

NAmap <- ggplot() + geom_polygon(data = usa, 
                                 aes(x=long, y = lat, group = group), 
                                 fill = "white", 
                                 color="black") +
    geom_polygon(data = canada, aes(x=long, y = lat, group = group), 
                 fill = "white", color="black") +
    coord_fixed(xlim = c(-95, -70),  ylim = c(33, 50), ratio = 1.2)
NAmap

range<-NAmap+ geom_point(data=inat_clean_corrected, aes(x=longitude, y=latitude), fill="hotpink",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "*Trillium grandiflorum* observations") +theme(title = ggtext::element_markdown())
infections<-NAmap+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), fill="yellowgreen",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "Infected *T. grandiflorum* observations") +theme(title = ggtext::element_markdown())
rangemap<-plot_grid(range,infections, labels=c("A", "B"))
ggsave2("phytoplasma_rangemap.png", rangemap, dpi=600, width=10, height=8)
```

sf version with ggplot
```{r}
st_layers("gadm41_CAN.gpkg")
canada <- read_sf("gadm41_CAN.gpkg", layer = "ADM_ADM_1") #https://inbo.github.io/tutorials/tutorials/spatial_standards_vector/

#canada <- read_sf("gadm41_CAN_1.shp")
#gadm41_CAN_1.shp
st_layers("gadm41_USA.gpkg")
us <- read_sf("gadm41_USA.gpkg", layer = "ADM_ADM_1")
plot(canada)
plot(us)
#b = st_buffer(canada[1,1], units::set_units(10, km)) #|> st_cast("LINESTRING")

#https://www.naturalearthdata.com/downloads/10m-physical-vectors/10m-lakes/
# Create temp files
temp2 <- tempfile()
# Download the zip file and save unzipped
# Unzip the contents of the temp and save unzipped content in 'temp2'
unzip(zipfile = "ne_10m_lakes.zip", exdir = temp2)
# Read the shapefile. Alternatively make an assignment, such as f<-sf::read_sf(your_SHP_file)
lakes<-sf::read_sf(temp2)

rangev2<-ggplot() + geom_sf(data = canada, fill="white") + geom_sf(data = us, fill="white")+geom_sf(data=lakes)+coord_sf(xlim = c(-95, -70),  ylim = c(33, 50))+ geom_point(data=inat_clean_corrected, aes(x=longitude, y=latitude), fill="hotpink",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "*Trillium grandiflorum* observations") +theme(title = ggtext::element_markdown()) 
#THIS MADE ME NOTICE A FEW OBSERVATIONS IN LAKE ONTARIO- FIND AND REMOVE!
#could use county map and colour by number of observations within a given county?

infectionsv2<-ggplot() + geom_sf(data = canada, fill="white") + geom_sf(data = us, fill="white")+geom_sf(data=lakes)+coord_sf(xlim = c(-95, -70),  ylim = c(33, 50))+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), fill="limegreen",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "Infected *T. grandiflorum* observations") +theme(title = ggtext::element_markdown())
rangemapv2<-plot_grid(rangev2,infectionsv2, labels=c("A", "B"))
ggsave2("phytoplasma_rangemapV2.png", rangemapv2, dpi=600, width=10, height=8)
```
```{r}
#https://www.naturalearthdata.com/downloads/10m-cultural-vectors/parks-and-protected-lands/
#https://www.naturalearthdata.com/downloads/10m-cultural-vectors/10m-urban-area/ 
```

Equal area projections
https://r.geocompx.org/reproj-geo-data.html
https://bbest.github.io/R-adv-spatial-lessons/tidy.html#challenge-project-recalculate-area
https://epsg.io/9822-method
https://www.andrewheiss.com/blog/2023/07/28/gradient-map-fills-r-sf/ !!!
```{r}
library(geosphere)
st_crs(us) # get CRS
st_crs(canada)
st_crs(lakes)

crs_usa = '+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs'
usa_geo = us %>%
  st_transform(st_crs("EPSG:9822"))
st_crs(us)
st_crs(usa_geo)
canada_geo = canada %>%
  st_transform(st_crs("EPSG:9822"))
lakes_geo = lakes %>%
  st_transform(st_crs("EPSG:9822"))
basemap<-canada_geo+usa_geo+lakes_geo

ggplot() + geom_sf(data = canada, fill="white") + geom_sf(data = us, fill="white")+geom_sf(data=lakes)+coord_sf(xlim = c(-95, -70),  ylim = c(33, 50))+ geom_point(data=inat_clean_corrected, aes(x=longitude, y=latitude), fill="hotpink",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "*Trillium grandiflorum* observations") +theme(title = ggtext::element_markdown()) 
#THIS MADE ME NOTICE A FEW OBSERVATIONS IN LAKE ONTARIO- FIND AND REMOVE!
#could use county map and colour by number of observations within a given county?
ggplot() + geom_sf(data = canada_geo, fill="white") + geom_sf(data = usa_geo, fill="white")+geom_sf(data=lakes_geo)+coord_sf(xlim = c(-95, -70),  ylim = c(33, 50), crs = "EPSG:9822")+ geom_point(data=inatinfec, aes(x=longitude, y=latitude), fill="limegreen",shape=21, size=1,alpha=0.3)+labs(x = "Longitude",y = "Latitude",title= "Infected *T. grandiflorum* observations") +theme(title = ggtext::element_markdown())

infection_coords_geo <- st_as_sf(x = inatinfec,                         
           coords = c("longitude", "latitude"),
           crs = "EPSG:9822")
observation_coords_geo <- st_as_sf(x = inat_clean_corrected,                         
           coords = c("longitude", "latitude"),
           crs = "EPSG:9822")
```
